---
title: "MM GFW VP & AFE Data Pull"
author: "Malachy McCaffrey"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Filter gridded vessel presence and apparent fishing effort data from 4Wings API for Orsted projects AOI between 2016 - 2026 then filter for only fishing vessels. Manipulate data into usable format, plot, and export as .csv

## Libraries

```{r}
library(gfwr)
library(tidyverse)
library(sf)
library(lubridate)
```

## Define key, date ranges, and spatial extent for data querying

```{r}

key <- Sys.getenv("GFW_TOKEN")

# API queries can only process one year of data at a time
# if request still times, out, split into smaller requests and bind results in R
# yyy-mm-dd format

start_15 <- '2015-01-01'
end_15 <- '2016-01-01'

start_16 <- '2016-01-01'
end_16 <- '2017-01-01'

start_17 <- '2017-01-01' # will be included
end_17 <- '2018-01-01' # will be excluded. search will be up to 2017-12-31

start_18 <- '2018-01-01'
end_18 <- '2019-01-01'

start_19 <- '2019-01-01'
end_19 <- '2020-01-01'

start_20 <- '2020-01-01'
end_20 <- '2021-01-01'

start_21 <- '2021-01-01'
end_21 <- '2022-01-01'

start_22 <- '2022-01-01'
end_22 <- '2023-01-01'

start_23 <- '2023-01-01'
end_23 <- '2024-01-01'

start_24 <- '2024-01-01'
end_24 <- '2025-01-01'

start_25 <- '2025-01-01'
end_25 <- '2026-01-01'

start_26 <- '2026-01-01'
end_26 <- '2027-01-01' # as of 07/14/2026
```

```{r}
# define sf object to pass to `region` argument

orsted_aoi <- st_read("shapefiles/Orsted_AOI.shp") |> 
  select(geometry)

# map aoi to make sure it was created correctly

ggplot(orsted_aoi) + geom_sf()
```

## Retrieve vessel presence data within aoi and date ranges grouped by vessel_id

```{r}
# 2016

vp_id_16 <- gfw_ais_presence(spatial_resolution = "HIGH", # 0.01 degree ~ 1 km
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", # associated with a single MMSI at a specific period of time. single vessel can have several IDs thru time
                               start_date = start_16, # date range must be 366 days or less
                               end_date = end_16,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2017

vp_id_17 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_17, 
                               end_date = end_17,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)


# 2018

vp_id_18 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_18, 
                               end_date = end_18,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2019

vp_id_19 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_19,
                               end_date = end_19,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2020

vp_id_20 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_20, 
                               end_date = end_20,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2021

vp_id_21 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_21, 
                               end_date = end_21,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2022

vp_id_22 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_22, 
                               end_date = end_22,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2023

vp_id_23 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_23, 
                               end_date = end_23,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2024

vp_id_24 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_24, 
                               end_date = end_24,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2025

vp_id_25 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_25, 
                               end_date = end_25,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2026

vp_id_26 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_26, 
                               end_date = end_26,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)
```

## Subset VP datasets to fishing vessel type

```{r}
# 2016

vp_fv_16 <- vp_id_16 |> 
  filter(`Vessel Type` == "FISHING")

# 2017

vp_fv_17 <- vp_id_17 |> 
  filter(`Vessel Type` == "FISHING")

# 2018

vp_fv_18 <- vp_id_18 |> 
  filter(`Vessel Type` == "FISHING")

# 2019

vp_fv_19 <- vp_id_19 |> 
  filter(`Vessel Type` == "FISHING")

# 2020

vp_fv_20 <- vp_id_20 |> 
  filter(`Vessel Type` == "FISHING")

# 2021

vp_fv_21 <- vp_id_21 |> 
  filter(`Vessel Type` == "FISHING")

# 2022

vp_fv_22 <- vp_id_22 |> 
  filter(`Vessel Type` == "FISHING")

# 2023

vp_fv_23 <- vp_id_23 |> 
  filter(`Vessel Type` == "FISHING")

# 2024

vp_fv_24 <- vp_id_24 |> 
  filter(`Vessel Type` == "FISHING")

# 2025

vp_fv_25 <- vp_id_25 |> 
  filter(`Vessel Type` == "FISHING")

# 2026

vp_fv_26 <- vp_id_26 |> 
  filter(`Vessel Type` == "FISHING")
```

## Retrieve apparent fishing effort data within aoi and date ranges grouped by vessel_id

```{r}
# 2016

afe_id_16 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_16, 
                               end_date = end_16,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2017

afe_id_17 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_17, 
                               end_date = end_17,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)


# 2018

afe_id_18 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_18, 
                               end_date = end_18,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2019

afe_id_19 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_19,
                               end_date = end_19,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2020

afe_id_20 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_20,
                               end_date = end_20,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2021

afe_id_21 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_21,
                               end_date = end_21,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2022

afe_id_22 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_22,
                               end_date = end_22,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2023

afe_id_23 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_23, 
                               end_date = end_23,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2024

afe_id_24 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_24, 
                               end_date = end_24,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2025

afe_id_25 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_25, 
                               end_date = end_25,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2026

afe_id_26 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_26,
                               end_date = end_26,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)
```

## Combine datasets

```{r}
# append rows
gfw_vp_id <- rbind(vp_id_16, vp_id_17, vp_id_18, vp_id_19, vp_id_20, vp_id_21, vp_id_22, vp_id_23, vp_id_24, vp_id_25, vp_id_26)

gfw_vp_fv <- rbind(vp_fv_16, vp_fv_17, vp_fv_18, vp_fv_19, vp_fv_20, vp_fv_21, vp_fv_22, vp_fv_23, vp_fv_24, vp_fv_25, vp_fv_26)

gfw_afe_id <- rbind(afe_id_16, afe_id_17, afe_id_18, afe_id_19, afe_id_20, afe_id_21, afe_id_22, afe_id_23, afe_id_24, afe_id_25, afe_id_26)
```

## create yyyy-mm column

```{r}
gfw_vp_id$YearMonth <- as.Date(format(gfw_vp_id$`Time Range`, "%Y-%m-01"))

gfw_vp_fv$YearMonth <- as.Date(format(gfw_vp_fv$`Time Range`, "%Y-%m-01"))

gfw_afe_id$YearMonth <- as.Date(format(gfw_afe_id$`Time Range`, "%Y-%m-01"))
```

## Create development stage column

```{r}
gfw_vp_id <- gfw_vp_id |>
  mutate(
    `Development Stage` = case_when(
      between(`YearMonth`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 44 months of pre-monitoring

      between(`YearMonth`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 43 months of pre-construction monitoring 

      between(`YearMonth`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 40 months of construction & operation across SFW, RWF, and SRW
    )
  )

gfw_vp_fv <- gfw_vp_fv |>
  mutate(
    `Development Stage` = case_when(
      between(`YearMonth`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 31 months of pre-monitoring

      between(`YearMonth`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 42 months of pre-construction monitoring 

      between(`YearMonth`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 37 months of construction & operation across SFW, RWF, and SRW
    )
  )

gfw_afe_id <- gfw_afe_id |>
  mutate(
    `Development Stage` = case_when(
      between(`YearMonth`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 31 months of pre-monitoring

      between(`YearMonth`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 42 months of pre-construction monitoring 

      between(`YearMonth`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 37 months of construction & operation across SFW, RWF, and SRW
    )
  )
```

## Chrs to factors

```{r}
gfw_vp_id$`Vessel Type` <- as.factor(gfw_vp_id$`Vessel Type`)
gfw_vp_id$`Vessel Name` <- as.factor(gfw_vp_id$`Vessel Name`)
gfw_vp_id$`Gear Type` <- as.factor(gfw_vp_id$`Gear Type`)
gfw_vp_id$`Development Stage` <- as.factor(gfw_vp_id$`Development Stage`)
gfw_vp_id$MMSI <- as.factor(gfw_vp_id$MMSI)
gfw_vp_id$Flag <-  as.factor(gfw_vp_id$Flag)
gfw_vp_id$`Vessel ID` <-  as.factor(gfw_vp_id$`Vessel ID`)

gfw_vp_fv$`Gear Type` <- as.factor(gfw_vp_fv$`Gear Type`)
gfw_vp_fv$`Vessel Name` <- as.factor(gfw_vp_fv$`Vessel Name`)
gfw_vp_fv$`Development Stage` <- as.factor(gfw_vp_fv$`Development Stage`)
gfw_vp_fv$MMSI <- as.factor(gfw_vp_fv$MMSI)
gfw_vp_fv$Flag <-  as.factor(gfw_vp_fv$Flag)
gfw_vp_fv$`Vessel ID` <-  as.factor(gfw_vp_fv$`Vessel ID`)

gfw_afe_id$`Gear Type` <- as.factor(gfw_afe_id$`Gear Type`)
gfw_afe_id$`Vessel Name` <- as.factor(gfw_afe_id$`Vessel Name`)
gfw_afe_id$MMSI <- as.factor(gfw_afe_id$MMSI)
gfw_afe_id$Flag <-  as.factor(gfw_afe_id$Flag)
gfw_afe_id$`Vessel ID` <-  as.factor(gfw_afe_id$`Vessel ID`)
```

## Sniff test

```{r}
# check data

summary(gfw_vp_id)
summary(gfw_vp_fv)
summary(gfw_afe_id) 
## zeros present in afe data
## Other, inconclusive, and passenger vessels in afe: 
# Hat Trick listed is a fishing yacht. keep
# Pasque is fishing vessel out of Falmouth according to Marine Traffic. keep
# Viking Star and Viking Starship are offshore fishing charters out of Montauk, NY. keep.
# Off The Grid is a fishing yacht... keep

#### imposing threshold: remove any records where fishing hours are less than .33 hours (20 mins) ####
```

## Remove vessels

```{r}
# list of orsted survey and safety vessels to remove for dev stages 2 & 3
# cross-referenced names and MMSIs in gfw_vp_id

rmOrsted <- c("ASHLEY ANN II","HEATHER LYNN","AMELIA ANNE","ERICA KNIGHT","MISTER G","MORE MISERY","SWEET MISERY","CAILYN & MAREN","HARVEST MOON","BARBARA ANN","BARBARA ANNE","FLEET KING","GABRIELLE ELIZABETH","JO-ANN V","MISTER MARCO","F/V NEW HORIZON","PROVIDER", "FLEET QUEEN", "TRIUNFO", "PAMELA ANN", "KAREN NICOLE", "ANDREA A", "MADISON KATE", "ITALIAN PRINCESS", "SEA WIND", "OCEAN QUEEN", "BRITTANY ERYN", "HUNTER", "CHATHAM", "PATIENCE", "MARINER", "F/V HARVESTER", "KATHRYN MARIE", "CELTIC")

# isolate orsted vessels

gfw_vp_orstedVessels <- gfw_vp_id |>
  filter(
    `Vessel Name` %in% rmOrsted &
    `Development Stage` %in% c("Stage 2", "Stage 3")) # 55969 obs of original 416533... 12.5% of data from all vessel AIS records

# remove orsted vessels

gfw_vp_fv_sub <- gfw_vp_fv |>
  filter(
    !(`Vessel Name` %in% rmOrsted &
      `Development Stage` %in% c("Stage 2", "Stage 3"))) # 88088 obs of 137539... 36% of data removed

gfw_afe_id_sub <- gfw_afe_id |> 
  filter(
    !(`Vessel Name` %in% rmOrsted &
      `Development Stage` %in% c("Stage 2", "Stage 3"))) # 20722 obs of 61637... 66% of data removed

# list of cargo vessels present in vp_fv and afe_id

rmCargo <- c("CYGNUS", "SUNNANVIK")

# remove cargo vessels

gfw_vp_fv_sub <- gfw_vp_fv_sub |> 
  filter(!`Vessel Name` %in% rmCargo) # 88035 obs

gfw_afe_id_sub <- gfw_afe_id_sub |> 
  filter(!`Vessel Name` %in% rmCargo) # 20696 obs
```

## Remove AFE records where fishing hours are less than .33 hours (20 mins)

```{r}
gfw_afe_id_sub <- gfw_afe_id_sub |> 
  filter(`Apparent Fishing Hours` > 0.32) # 19068 obs
```

## remove unwanted cols

```{r}
# remove unwanted cols

gfw_vp_orstedVessels <- gfw_vp_orstedVessels[, -c(7,8,12:15)]
gfw_vp_fv_sub <- gfw_vp_fv_sub[,-c(7,8,12:15)]
gfw_afe_id_sub <- gfw_afe_id_sub[,-c(7,8,12:15)]
```

## filter vp_fv and afe data sets by development stage

```{r}
# stage 1 - pre-monitoring

gfw_vp1_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 1")

gfw_afe1_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 1")

# stage 2 - pre-construction monitoring

gfw_vp2_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 2")

gfw_afe2_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 2")

# stage 3 - construction and operation

gfw_vp3_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 3")

gfw_afe3_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 3")
```

## more sniff tests - visualization checks

```{r}
# histograms

gfw_afe_id_sub |> 
  ggplot(aes(x=`Apparent Fishing Hours`)) + geom_histogram(binwidth = 0.5) + theme_bw()

gfw_vp_fv_sub |> 
  ggplot(aes(x=`Vessel Presence Hours`)) + geom_histogram(binwidth = 1) + theme_bw()

# how to handle one-inflation in data sets?
```

```{r}
# check dist of records by month

gfw_vp_fv_sub |> 
  ggplot(aes(x = YearMonth)) +
  geom_bar(fill = "steelblue") +
  theme_bw()

gfw_afe_id_sub |> 
  ggplot(aes(x = YearMonth)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # why sudden jump in mid/late 2024 and 2025?

# check by dev stage

gfw_vp_fv_sub |> 
  ggplot(aes(x = `Development Stage`)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # steady increase a result of just more general fishing activity over time?

gfw_afe_id_sub |> 
  ggplot(aes(x = `Development Stage`)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # BIG jump in dev stage 3
```

```{r}
# check dist of unique vessel IDs by month

IDs_perMonth <- gfw_vp_fv_sub |> 
  group_by(YearMonth) |> 
  summarise(unique_id_count = n_distinct(`Vessel ID`))

IDs_perMonth |> 
  ggplot(aes(x = as.factor(YearMonth), y = unique_id_count)) +
  geom_col(fill = "steelblue") +
  theme_bw() # trend seems consistent over time... not seeing any big spikes in later years that should indicate more vessels fishing in the area

# check by dev stage

IDs_perDev <- gfw_vp_fv_sub |> 
  group_by(`Development Stage`) |> 
  summarise(unique_id_count = n_distinct(`Vessel ID`))

IDs_perDev |> 
  ggplot(aes(x = as.factor(`Development Stage`), y = unique_id_count)) +
  geom_col(fill = "steelblue") +
  theme_bw() # slight decrease in no. of vessels from stage 1 to 2 then small rebound in 3
```

## Export data sets

```{r}
# all vp

write_csv(gfw_vp_id, "processed/gfw_vp_id.csv")

# orsted vessels 

write_csv(gfw_vp_orstedVessels, "processed/gfw_vp_orstedVessels.csv")

# vp_fv

write_csv(gfw_vp_fv, "processed/gfw_vp_fv.csv")
write_csv(gfw_vp_fv_sub, "processed/gfw_vp_fv_sub.csv")
write_csv(gfw_vp1_fv_sub, "processed/gfw_vp1_fv_sub.csv")
write_csv(gfw_vp2_fv_sub, "processed/gfw_vp2_fv_sub.csv")
write_csv(gfw_vp3_fv_sub, "processed/gfw_vp3_fv_sub.csv")

# afe_id

write_csv(gfw_afe_id, "gfw_afe_id.csv")
write_csv(gfw_afe_id_sub, "processed/gfw_afe_id_sub.csv")
write_csv(gfw_afe1_id_sub, "processed/gfw_afe1_id_sub.csv")
write_csv(gfw_afe2_id_sub, "processed/gfw_afe2_id_sub.csv")
write_csv(gfw_afe3_id_sub, "processed/gfw_afe3_id_sub.csv")
```