---
title: "MM GFW 0.01° Fishnet Grid (UTM 19N)"
author: "Malachy McCaffrey"
date: today
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Reproduce Global Fishing Watch's 0.01° grid for the vessel presence (VP) and
apparent fishing effort (AFE) records pulled in `gfw_vp_afe_datapull.qmd`,
reproject it to WGS 84 / UTM zone 19N (EPSG:32619), and attach the data to it
for aggregation and spatial analysis.

### Data status: aggregates describe the raw GFW data

The aggregate tables built here describe the **raw, unfiltered** GFW records in
`data/raw/gfw/gfw_vp_id.csv` and `gfw_afe_id.csv`. They include every vessel
type and apply no vessel removals or fishing-hour threshold. They are **not**
fishing-vessel totals and should not be analysed as such.

The data have not yet been re-processed because GFW's vessel classifications
produce both false negatives and false positives:

- **False negatives:** fishing vessels typed as `OTHER`, `GEAR`, or
  `PASSENGER` (e.g. FV JULIE G, ONWA33NT), so filtering on
  `Vessel Type == "FISHING"` drops real fishing activity.
- **False positives:** non-fishing vessels typed as `FISHING` (e.g. CYGNUS), so
  the same filter keeps non-fishing activity.

Both problems affect VP and AFE. Once vessel classification is resolved,
re-run the aggregation on the re-processed data; the grid itself does not
change, because it is built from the cells present in the raw VP data.

## Plan

1. **Snap coordinates to integer cell indices.** GFW `Lat`/`Lon` mark the
   **lower-left corner** of each 0.01° cell. Values carry float32 noise
   (e.g. `-71.209999`), so snap with `round(x / 0.01)` -- never `floor()`, which
   drops values like `41.0699999` into the wrong cell. The integer pair
   `cell_id = "ix_iy"` is the join key from here on; floating-point
   coordinates are never compared.
2. **Grid = the distinct cells in `gfw_vp_id`** (3,376 cells). Every AFE cell
   is already among them.
3. **Build each cell from its lower-left corner** in WGS 84, spanning
   `Lon`–`Lon + 0.01` and `Lat`–`Lat + 0.01`. Corners come from integer
   arithmetic so adjacent cells share bit-identical vertices.
4. **Reproject the polygons to EPSG:32619.** Each cell becomes a parallelogram
   of about 838 m E–W by 1,111 m N–S, rotated about 1.6° by grid convergence.
   Do not build a new square UTM fishnet or resample a raster, either of which
   would redistribute hours between cells.
5. **Add cell attributes:** `area_km2` (from UTM, 0.929–0.937 km²),
   `aoi_frac`, and `aoi_class` (inside / edge / outside).
6. **Grid-cell decisions** (see below): keep the 52 cells outside the AOI;
   exclude the 51 AOI-touching cells that have no records.
7. **Attach the raw data with a table join on `cell_id`**, aggregate per
   cell × stage and per cell × month, and fill zeros across the 3,376-cell
   grid. These are raw-data aggregates (all vessel types); see *Data status*.
8. **Neighbours** from queen contiguity on the UTM grid.

### Grid-cell decisions

Under the lower-left anchor, two sets of cells sit at the AOI boundary:

- **52 cells outside the AOI -- kept.** All lie along the north and east
  edges, up to 568 m outside the boundary. They are cells GFW returned for the
  AOI query and hold real records: 3,063 VP rows (0.69% of rows, 0.44% of
  hours) and 295 AFE rows in 38 of the 52 cells. They are flagged
  `aoi_class == "outside"` so they can be dropped in a sensitivity check.
- **51 AOI-touching cells with no records -- excluded.** All lie along the
  south and west edges. No VP or AFE record has ever been returned under their
  keys, so they are not part of the grid and are not zero-filled.

The mirror-image pattern (extra cells to the NE, missing cells to the SW) is
consistent with GFW selecting cells for a `USER_SHAPEFILE` region by a rule
that differs from the lower-left labelling. Diagnostics are in
`r/scratchpad/gridcell_decision.R`.

### Why a table join, not a spatial join

Each record already names its cell, so the join only has to read the label.
With lower-left anchoring every point sits exactly on a cell corner shared by
four cells, and float noise nudges it to an arbitrary side of the edge. On a
20,000-row sample of `gfw_vp_id`:

- `st_intersects` (WGS 84) matched most points to 2 or 4 cells, turning 20,000
  rows into 44,183 and inflating VP hours 2.19×.
- `st_within` (WGS 84) matched none of the 20,000 points.
- `st_intersects` after reprojecting to UTM put 24,197 rows in the wrong cell.

None of these fail loudly. The table join is exact, one-to-one, checkable with
`anti_join()`, and identical in any language. Spatial joins are reserved for
linking the grid to layers that do not share the key (lease areas, turbines,
the AOI), done once at the cell level as in the `aoi_frac` step.

## Libraries

```{r}
library(tidyverse)
library(sf)
library(spdep)
```

## Parameters

```{r}
res     <- 0.01    # GFW "HIGH" spatial resolution, degrees
crs_utm <- 32619   # WGS 84 / UTM zone 19N
```

## Read raw data and snap to cell indices

```{r}
# GFW Lat/Lon = lower-left corner of a 0.01 degree cell. Values carry float32
# noise (e.g. -71.209999), so round() -- never floor() -- to the lattice.
read_gfw <- function(path) {
  read_csv(path, show_col_types = FALSE,
           col_types = cols(MMSI = col_character())) |>
    mutate(ix = as.integer(round(Lon / res)),
           iy = as.integer(round(Lat / res)),
           cell_id = sprintf("%d_%d", ix, iy))
}

vp  <- read_gfw("data/raw/gfw/gfw_vp_id.csv")
afe <- read_gfw("data/raw/gfw/gfw_afe_id.csv")

# every coordinate must sit on the 0.01 lattice (observed max error: 4e-4 cells)
snap_err <- with(bind_rows(vp, afe), max(abs(Lon / res - ix), abs(Lat / res - iy)))
stopifnot("coordinates are not on the 0.01 degree lattice" = snap_err < 0.01)
```

## Grid cells from the VP data

```{r}
cells <- distinct(vp, cell_id, ix, iy) # 3376 cells

stopifnot("AFE has cells absent from the VP grid" =
            all(afe$cell_id %in% cells$cell_id))
```

## Build cell polygons from the lower-left corner

```{r}
# Corners come from integer arithmetic so adjacent cells share bit-identical
# vertices, which keeps topology exact after reprojection.
cell_poly <- function(ix, iy) {
  x <- (ix + c(0L, 1L, 1L, 0L, 0L)) / 100
  y <- (iy + c(0L, 0L, 1L, 1L, 0L)) / 100
  st_polygon(list(cbind(x, y)))
}

grid_wgs <- st_sf(cells,
                  geometry = st_sfc(map2(cells$ix, cells$iy, cell_poly),
                                    crs = 4326))
```

## Reproject and add cell attributes

```{r}
aoi_utm <- st_read("data/sf/aoi/Orsted_AOI.shp", quiet = TRUE) |>
  st_union() |>
  st_transform(crs_utm)

grid_utm <- st_transform(grid_wgs, crs_utm) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) # 0.929-0.937 km2

# fraction of each cell inside the AOI; cells outside it get 0
in_aoi <- st_intersection(select(grid_utm, cell_id), aoi_utm)

grid_utm <- grid_utm |>
  left_join(tibble(cell_id = in_aoi$cell_id,
                   aoi_m2  = as.numeric(st_area(in_aoi))),
            by = "cell_id") |>
  mutate(aoi_frac  = coalesce(aoi_m2 / (area_km2 * 1e6), 0),
         aoi_class = factor(case_when(aoi_frac == 0   ~ "outside",
                                      aoi_frac < 0.99 ~ "edge",
                                      TRUE            ~ "inside"),
                            levels = c("inside", "edge", "outside"))) |>
  select(-aoi_m2)

count(st_drop_geometry(grid_utm), aoi_class) # 3118 inside, 206 edge, 52 outside
```

## Grid-cell decision checks

```{r}
# Kept: the 52 data cells outside the AOI stay in the grid, flagged "outside".
stopifnot("expected 52 cells outside the AOI" =
            sum(grid_utm$aoi_class == "outside") == 52)

# Excluded: AOI-touching cells with no records are not added to the grid.
# Listed here only to document the decision.
lattice <- expand_grid(ix = seq(min(cells$ix) - 1L, max(cells$ix) + 1L),
                       iy = seq(min(cells$iy) - 1L, max(cells$iy) + 1L)) |>
  mutate(cell_id = sprintf("%d_%d", ix, iy)) |>
  filter(!cell_id %in% cells$cell_id)

empty_utm <- st_sf(lattice,
                   geometry = st_sfc(map2(lattice$ix, lattice$iy, cell_poly),
                                     crs = 4326)) |>
  st_transform(crs_utm)

empty_utm <- empty_utm[lengths(st_intersects(empty_utm, aoi_utm)) > 0, ]

nrow(empty_utm) # 51 cells excluded

stopifnot("excluded empty cells leaked into the grid" =
            !any(empty_utm$cell_id %in% grid_utm$cell_id))
```

## Map check

```{r}
ggplot() +
  geom_sf(data = grid_utm, aes(fill = aoi_class), color = "white", linewidth = 0.05) +
  geom_sf(data = empty_utm, fill = NA, color = "black", linewidth = 0.3, linetype = "22") +
  geom_sf(data = aoi_utm, fill = NA, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(inside = "#d9d8d2", edge = "#2a78d6", outside = "#eb6834")) +
  labs(fill = NULL,
       caption = "Dashed: 51 AOI-touching cells with no records (excluded).") +
  theme_minimal()
```

## Attach raw data with a table join and aggregate

These aggregates describe the **raw GFW data** -- all vessel types, no vessel
removals, no fishing-hour threshold. They are not fishing-vessel totals; see
*Data status* above.

```{r}
# The raw files carry only `Time Range`, so derive `Year Month` and
# `Development Stage`. Stage boundaries mirror gfw_vp_afe_datapull.qmd --
# keep the two in sync.
add_time_fields <- function(df) {
  df |>
    mutate(`Year Month` = floor_date(`Time Range`, "month"),
           `Development Stage` = case_when(
             between(`Year Month`, as.Date("2016-01-01"), as.Date("2019-08-01")) ~ "Stage 1",
             between(`Year Month`, as.Date("2019-09-01"), as.Date("2023-03-01")) ~ "Stage 2",
             between(`Year Month`, as.Date("2023-04-01"), as.Date("2026-09-01")) ~ "Stage 3"))
}

vp_raw  <- add_time_fields(vp)
afe_raw <- add_time_fields(afe)

# every record must land in a grid cell -- both should return zero rows
grid_ids <- st_drop_geometry(grid_utm) |> select(cell_id)

stopifnot(
  "VP records with no grid cell"  = nrow(anti_join(vp_raw,  grid_ids, by = "cell_id")) == 0,
  "AFE records with no grid cell" = nrow(anti_join(afe_raw, grid_ids, by = "cell_id")) == 0
)

# stages run Jan 2016 - Sep 2026, covering the full pull; every record needs one
stopifnot(
  "VP records with no development stage"  = !anyNA(vp_raw$`Development Stage`),
  "AFE records with no development stage" = !anyNA(afe_raw$`Development Stage`)
)

count(vp_raw, `Development Stage`)
```

```{r}
# cell x development stage, zero-filled across the 3376-cell grid
# counting unit is MMSI, not Vessel ID (see references/data_dictionary.md)

vp_cell_stage <- vp_raw |>
  group_by(cell_id, `Development Stage`) |>
  summarise(vp_hours = sum(`Vessel Presence Hours`),
            n_mmsi   = n_distinct(MMSI),
            .groups  = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Development Stage`,
           fill = list(vp_hours = 0, n_mmsi = 0L))

afe_cell_stage <- afe_raw |>
  group_by(cell_id, `Development Stage`) |>
  summarise(afe_hours = sum(`Apparent Fishing Hours`),
            n_mmsi    = n_distinct(MMSI),
            .groups   = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Development Stage`,
           fill = list(afe_hours = 0, n_mmsi = 0L))
```

```{r}
# cell x month, zero-filled across the grid and every month in the record

months <- seq(min(vp_raw$`Year Month`), max(vp_raw$`Year Month`), by = "month")

vp_cell_month <- vp_raw |>
  group_by(cell_id, `Year Month`) |>
  summarise(vp_hours = sum(`Vessel Presence Hours`),
            n_mmsi   = n_distinct(MMSI),
            .groups  = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Year Month` = months,
           fill = list(vp_hours = 0, n_mmsi = 0L))

afe_cell_month <- afe_raw |>
  group_by(cell_id, `Year Month`) |>
  summarise(afe_hours = sum(`Apparent Fishing Hours`),
            n_mmsi    = n_distinct(MMSI),
            .groups   = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Year Month` = months,
           fill = list(afe_hours = 0, n_mmsi = 0L))

stopifnot(nrow(vp_cell_month) == nrow(grid_utm) * length(months))
```

```{r}
# spatial versions with area-normalised density, since cell area varies ~1%

vp_stage_sf <- grid_utm |>
  left_join(vp_cell_stage, by = "cell_id") |>
  mutate(vp_hours_km2 = vp_hours / area_km2)

afe_stage_sf <- grid_utm |>
  left_join(afe_cell_stage, by = "cell_id") |>
  mutate(afe_hours_km2 = afe_hours / area_km2)
```

## Neighbours

```{r}
# Cells are ~838 x 1,111 m, not square: a distance band would link E-W
# neighbours but skip N-S ones, so use contiguity.
nb <- poly2nb(grid_utm, queen = TRUE, snap = 0)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

summary(card(nb)) # mean ~7.8 links, no islands
stopifnot("grid has cells with no neighbours" = all(card(nb) > 0))
```

## Export

```{r}
# grid (supersedes archive/fishnet/Orsted_sqGrid_utm19n.shp, which was built
# treating Lat/Lon as centroids)
st_write(grid_utm, "data/sf/fishnet/gfw_grid_utm19n.gpkg", delete_dsn = TRUE)

# raw-data aggregates (all vessel types); file names keep the raw "_id" suffix
# so they are not mistaken for fishing-vessel totals
write_csv(vp_cell_stage,  "data/interim/gfw/gfw_vp_id_cell_stage.csv")
write_csv(afe_cell_stage, "data/interim/gfw/gfw_afe_id_cell_stage.csv")
write_csv(vp_cell_month,  "data/interim/gfw/gfw_vp_id_cell_month.csv")
write_csv(afe_cell_month, "data/interim/gfw/gfw_afe_id_cell_month.csv")
```
