---
title: "MM GFW VP & AFE Data Pull"
author: "Malachy McCaffrey"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Filter gridded vessel presence and apparent fishing effort data from 4Wings API for Orsted projects AOI between 2016 - 2026 then filter for only fishing vessels. Manipulate data into usable format, plot, and export as .csv

## Libraries

```{r}
library(gfwr)
library(tidyverse)
library(sf)
library(lubridate)
```

## Define key, date ranges, and spatial extent for data querying

```{r}

key <- Sys.getenv("GFW_TOKEN")

# API queries can only process one year of data at a time
# if request still times, out, split into smaller requests and bind results in R
# yyy-mm-dd format

start_15 <- '2015-01-01'
end_15 <- '2016-01-01'

start_16 <- '2016-01-01'
end_16 <- '2017-01-01'

start_17 <- '2017-01-01' # will be included
end_17 <- '2018-01-01' # will be excluded. search will be up to 2017-12-31

start_18 <- '2018-01-01'
end_18 <- '2019-01-01'

start_19 <- '2019-01-01'
end_19 <- '2020-01-01'

start_20 <- '2020-01-01'
end_20 <- '2021-01-01'

start_21 <- '2021-01-01'
end_21 <- '2022-01-01'

start_22 <- '2022-01-01'
end_22 <- '2023-01-01'

start_23 <- '2023-01-01'
end_23 <- '2024-01-01'

start_24 <- '2024-01-01'
end_24 <- '2025-01-01'

start_25 <- '2025-01-01'
end_25 <- '2026-01-01'

start_26 <- '2026-01-01'
end_26 <- '2027-01-01' # as of 07/14/2026
```

```{r}
# define sf object to pass to `region` argument

orsted_aoi <- st_read("data/shp/aoi/Orsted_AOI.shp") |> 
  select(geometry)

# map aoi to make sure it was created correctly

ggplot(orsted_aoi) + geom_sf()
```

## Retrieve vessel presence data within aoi and date ranges grouped by vessel_id

```{r}
# 2016

vp_id_16 <- gfw_ais_presence(spatial_resolution = "HIGH", # 0.01 degree ~ 1 km
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", # associated with a single MMSI at a specific period of time. single vessel can have several IDs thru time
                               start_date = start_16, # date range must be 366 days or less
                               end_date = end_16,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2017

vp_id_17 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_17, 
                               end_date = end_17,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)


# 2018

vp_id_18 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_18, 
                               end_date = end_18,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2019

vp_id_19 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_19,
                               end_date = end_19,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2020

vp_id_20 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_20, 
                               end_date = end_20,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2021

vp_id_21 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_21, 
                               end_date = end_21,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2022

vp_id_22 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_22, 
                               end_date = end_22,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2023

vp_id_23 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_23, 
                               end_date = end_23,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2024

vp_id_24 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_24, 
                               end_date = end_24,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2025

vp_id_25 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_25, 
                               end_date = end_25,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2026

vp_id_26 <- gfw_ais_presence(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_26, 
                               end_date = end_26,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)
```

## Subset VP datasets to fishing vessel type

```{r}
# 2016

vp_fv_16 <- vp_id_16 |> 
  filter(`Vessel Type` == "FISHING")

# 2017

vp_fv_17 <- vp_id_17 |> 
  filter(`Vessel Type` == "FISHING")

# 2018

vp_fv_18 <- vp_id_18 |> 
  filter(`Vessel Type` == "FISHING")

# 2019

vp_fv_19 <- vp_id_19 |> 
  filter(`Vessel Type` == "FISHING")

# 2020

vp_fv_20 <- vp_id_20 |> 
  filter(`Vessel Type` == "FISHING")

# 2021

vp_fv_21 <- vp_id_21 |> 
  filter(`Vessel Type` == "FISHING")

# 2022

vp_fv_22 <- vp_id_22 |> 
  filter(`Vessel Type` == "FISHING")

# 2023

vp_fv_23 <- vp_id_23 |> 
  filter(`Vessel Type` == "FISHING")

# 2024

vp_fv_24 <- vp_id_24 |> 
  filter(`Vessel Type` == "FISHING")

# 2025

vp_fv_25 <- vp_id_25 |> 
  filter(`Vessel Type` == "FISHING")

# 2026

vp_fv_26 <- vp_id_26 |> 
  filter(`Vessel Type` == "FISHING")
```

## Retrieve apparent fishing effort data within aoi and date ranges grouped by vessel_id

```{r}
# 2016

afe_id_16 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_16, 
                               end_date = end_16,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2017

afe_id_17 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_17, 
                               end_date = end_17,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)


# 2018

afe_id_18 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_18, 
                               end_date = end_18,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2019

afe_id_19 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_19,
                               end_date = end_19,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2020

afe_id_20 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_20,
                               end_date = end_20,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2021

afe_id_21 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_21,
                               end_date = end_21,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2022

afe_id_22 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_22,
                               end_date = end_22,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2023

afe_id_23 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID", 
                               start_date = start_23, 
                               end_date = end_23,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2024

afe_id_24 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH", 
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_24, 
                               end_date = end_24,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2025

afe_id_25 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_25, 
                               end_date = end_25,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)

# 2026

afe_id_26 <- gfw_ais_fishing_hours(spatial_resolution = "HIGH",
                               temporal_resolution = "DAILY",
                               group_by = "VESSEL_ID",
                               start_date = start_26,
                               end_date = end_26,
                               region_source = "USER_SHAPEFILE",
                               region = orsted_aoi)
```

## Combine datasets

```{r}
# append rows
gfw_vp_id <- rbind(vp_id_16, vp_id_17, vp_id_18, vp_id_19, vp_id_20, vp_id_21, vp_id_22, vp_id_23, vp_id_24, vp_id_25, vp_id_26)

gfw_vp_fv <- rbind(vp_fv_16, vp_fv_17, vp_fv_18, vp_fv_19, vp_fv_20, vp_fv_21, vp_fv_22, vp_fv_23, vp_fv_24, vp_fv_25, vp_fv_26)

gfw_afe_id <- rbind(afe_id_16, afe_id_17, afe_id_18, afe_id_19, afe_id_20, afe_id_21, afe_id_22, afe_id_23, afe_id_24, afe_id_25, afe_id_26)
```

## create yyyy-mm column

```{r}
gfw_vp_id$'Year Month' <- as.Date(format(gfw_vp_id$`Time Range`, "%Y-%m-01"))

gfw_vp_fv$'Year Month' <- as.Date(format(gfw_vp_fv$`Time Range`, "%Y-%m-01"))

gfw_afe_id$'Year Month' <- as.Date(format(gfw_afe_id$`Time Range`, "%Y-%m-01"))
```

## Create development stage column

```{r}
gfw_vp_id <- gfw_vp_id |>
  mutate(
    `Development Stage` = case_when(
      between(`Year Month`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 44 months of pre-monitoring

      between(`Year Month`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 43 months of pre-construction monitoring 

      between(`Year Month`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 40 months of construction & operation across SFW, RWF, and SRW
    )
  )

gfw_vp_fv <- gfw_vp_fv |>
  mutate(
    `Development Stage` = case_when(
      between(`Year Month`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 44 months of pre-monitoring

      between(`Year Month`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 43 months of pre-construction monitoring

      between(`Year Month`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 40 months of construction & operation across SFW, RWF, and SRW
    )
  )

gfw_afe_id <- gfw_afe_id |>
  mutate(
    `Development Stage` = case_when(
      between(`Year Month`,
              as.Date("2016-01-01"),
              as.Date("2019-08-01")) ~ "Stage 1", # 31 months of pre-monitoring

      between(`Year Month`,
              as.Date("2019-09-01"),
              as.Date("2023-03-01")) ~ "Stage 2", # protected species surveys began in 9/19. 42 months of pre-construction monitoring 

      between(`Year Month`,
              as.Date("2023-04-01"),
              as.Date("2026-07-17")) ~ "Stage 3" # sea bed prep for SFW began in 04/23. 37 months of construction & operation across SFW, RWF, and SRW
    )
  )
```

## Chrs to factors

```{r}
gfw_vp_id$`Vessel Type` <- as.factor(gfw_vp_id$`Vessel Type`)
gfw_vp_id$`Vessel Name` <- as.factor(gfw_vp_id$`Vessel Name`)
gfw_vp_id$`Gear Type` <- as.factor(gfw_vp_id$`Gear Type`)
gfw_vp_id$`Development Stage` <- as.factor(gfw_vp_id$`Development Stage`)
gfw_vp_id$MMSI <- as.factor(gfw_vp_id$MMSI)
gfw_vp_id$Flag <-  as.factor(gfw_vp_id$Flag)
gfw_vp_id$`Vessel ID` <-  as.factor(gfw_vp_id$`Vessel ID`)

gfw_vp_fv$`Gear Type` <- as.factor(gfw_vp_fv$`Gear Type`)
gfw_vp_fv$`Vessel Name` <- as.factor(gfw_vp_fv$`Vessel Name`)
gfw_vp_fv$`Development Stage` <- as.factor(gfw_vp_fv$`Development Stage`)
gfw_vp_fv$MMSI <- as.factor(gfw_vp_fv$MMSI)
gfw_vp_fv$Flag <-  as.factor(gfw_vp_fv$Flag)
gfw_vp_fv$`Vessel ID` <-  as.factor(gfw_vp_fv$`Vessel ID`)

gfw_afe_id$`Gear Type` <- as.factor(gfw_afe_id$`Gear Type`)
gfw_afe_id$`Vessel Name` <- as.factor(gfw_afe_id$`Vessel Name`)
gfw_afe_id$MMSI <- as.factor(gfw_afe_id$MMSI)
gfw_afe_id$Flag <-  as.factor(gfw_afe_id$Flag)
gfw_afe_id$`Vessel ID` <-  as.factor(gfw_afe_id$`Vessel ID`)
```

## Sniff test

```{r}
# check data

summary(gfw_vp_id)
summary(gfw_vp_fv)
summary(gfw_afe_id) 
## zeros present in afe data
## Other, inconclusive, and passenger vessels in afe: 
# Hat Trick listed is a fishing yacht. keep
# Pasque is fishing vessel out of Falmouth according to Marine Traffic. keep
# Viking Star and Viking Starship are offshore fishing charters out of Montauk, NY. keep.
# Off The Grid is a fishing yacht... keep

#### imposing threshold: remove any records where fishing hours are less than .33 hours (20 mins) ####
```

## Remove vessels

```{r}
# Orsted survey and safety vessels, plus non-fishing vessels, to remove.
#
# MATCHED ON MMSI, NOT VESSEL NAME. Name matching cannot reliably remove a
# vessel, because one hull emits several self-reported name strings. It failed
# in both directions:
#
#   - Leaked. "CAILYN & MAREN" and "CAILYN MAREN" are the same MMSI
#     (338389953); "F/V HARVESTER" and "HARVESTER" are both 367336020. The
#     name-based filter removed only the exact spellings listed and left the
#     variants in, worth ~260 vessel-presence hours in stage 3 alone.
#   - Over-matched. "TRADITION" spans six MMSIs across two vessel types, only
#     one of which is a charter, so any name rule would delete five unrelated
#     hulls. Likewise "KATHY MARIE" (366758680) is a different vessel from
#     "KATHRYN MARIE" (367387310), which is the one on the list.
#
# The vessel list is read from the running list rather than hard-coded here, so
# adding a vessel means editing the CSV, not this script.

rmPath <- "references/vessel_removals.csv"

rmVessels <- read_csv(rmPath, col_types = cols(.default = col_character()))

# --- Schema validation ---------------------------------------------------
# The removal list is the single most consequential judgement in this analysis,
# so it is validated on read rather than trusted. A typo, a truncated MMSI or a
# duplicated row would otherwise fail silently or, worse, not at all.

stopifnot(
  "removal list is missing required columns" =
    all(c("Vessel_Name", "MMSI", "Scope", "Identified_By",
          "Confidence", "Date_Added") %in% names(rmVessels))
)

rmProblems <- c(
  if (any(is.na(rmVessels$Vessel_Name) | rmVessels$Vessel_Name == ""))
    "blank Vessel_Name",
  if (any(!grepl("^[0-9]{9}$", rmVessels$MMSI)))
    paste0("MMSI is not 9 digits: ",
           paste(rmVessels$Vessel_Name[!grepl("^[0-9]{9}$", rmVessels$MMSI)],
                 collapse = ", ")),
  if (any(duplicated(rmVessels$MMSI)))
    paste0("duplicate MMSI: ",
           paste(unique(rmVessels$MMSI[duplicated(rmVessels$MMSI)]),
                 collapse = ", ")),
  if (!all(rmVessels$Scope %in% c("stages_2_3", "all_stages")))
    paste0("unknown Scope value: ",
           paste(setdiff(rmVessels$Scope, c("stages_2_3", "all_stages")),
                 collapse = ", ")),
  if (!all(rmVessels$Confidence %in% c("confirmed", "probable", "possible")))
    paste0("unknown Confidence value: ",
           paste(setdiff(rmVessels$Confidence,
                         c("confirmed", "probable", "possible")),
                 collapse = ", "))
)

if (length(rmProblems) > 0) {
  stop("Invalid ", rmPath, ":\n  - ", paste(rmProblems, collapse = "\n  - "),
       call. = FALSE)
}

# --- Derive the two removal sets from Scope ------------------------------
# Nothing vessel-specific is hard-coded here: adding a vessel, or changing how
# it is treated, is a one-row edit to the CSV.
#
# stages_2_3 -- chartered working fishing hulls. Their stage 1 records are real
#   fishing activity and are kept; only stages 2-3 project work is dropped.
# all_stages -- not fishing vessels at all, so there is no baseline to keep.
#   Currently just NOAA Gloria Michelle, a NOAA Fisheries research vessel that
#   GFW classes FISHING / TRAWLERS, inflating the mobile gear class.

rmCharter   <- rmVessels |> filter(Scope == "stages_2_3") |> pull(MMSI)
rmAllStages <- rmVessels |> filter(Scope == "all_stages") |> pull(MMSI)

# MMSI was coerced to a factor above, so compare on as.character().

# isolate orsted vessels

gfw_vp_orstedVessels <- gfw_vp_id |>
  filter(
    as.character(MMSI) %in% rmCharter &
    `Development Stage` %in% c("Stage 2", "Stage 3"))

# remove orsted charter vessels (stages 2-3) and non-fishing vessels (all stages)

gfw_vp_fv_sub <- gfw_vp_fv |>
  filter(
    !(as.character(MMSI) %in% rmCharter &
      `Development Stage` %in% c("Stage 2", "Stage 3")),
    !as.character(MMSI) %in% rmAllStages)

gfw_afe_id_sub <- gfw_afe_id |>
  filter(
    !(as.character(MMSI) %in% rmCharter &
      `Development Stage` %in% c("Stage 2", "Stage 3")),
    !as.character(MMSI) %in% rmAllStages)

# Report what was removed, so a re-run surfaces changes rather than hiding them.
message("Removed ", length(rmCharter), " charter vessels (stages 2-3) and ",
        length(rmAllStages), " non-fishing vessels (all stages). ",
        "vp_fv: ", nrow(gfw_vp_fv), " -> ", nrow(gfw_vp_fv_sub),
        " | afe: ", nrow(gfw_afe_id), " -> ", nrow(gfw_afe_id_sub))

# Cargo vessels that GFW misclassified as FISHING, removed from every stage.
#
# Also matched on MMSI, for the same reason as above: "CYGNUS" alone spans five
# MMSIs across three vessel types (FISHING, CARGO and PASSENGER), so a name
# rule removes four unrelated hulls along with the intended one.
#
# These two are the hulls the original name-based filter actually caught --
# both Norwegian-flagged and typed FISHING / INCONCLUSIVE by GFW. Confirmed by
# row count: they account for exactly the 53 vessel-presence and 26 apparent-
# effort records the original inline comments recorded.
rmCargo <- c("257394000",   # CYGNUS
             "259286000")   # SUNNANVIK

gfw_vp_fv_sub <- gfw_vp_fv_sub |>
  filter(!as.character(MMSI) %in% rmCargo)

gfw_afe_id_sub <- gfw_afe_id_sub |>
  filter(!as.character(MMSI) %in% rmCargo)
```

## Remove AFE records where fishing hours are less than .33 hours (20 mins)

```{r}
gfw_afe_id_sub <- gfw_afe_id_sub |> 
  filter(`Apparent Fishing Hours` > 0.32) # 17311... 1595 obs removed
```

## remove unwanted cols

```{r}
# remove unwanted cols

gfw_vp_orstedVessels <- gfw_vp_orstedVessels[, -c(7,8,12:15)]
gfw_vp_fv_sub <- gfw_vp_fv_sub[,-c(7,8,12:15)]
gfw_afe_id_sub <- gfw_afe_id_sub[,-c(7,8,12:15)]
```

## filter vp_fv and afe data sets by development stage

```{r}
# stage 1 - pre-monitoring

gfw_vp1_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 1")

gfw_afe1_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 1")

# stage 2 - pre-construction monitoring

gfw_vp2_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 2")

gfw_afe2_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 2")

# stage 3 - construction and operation

gfw_vp3_fv_sub <- gfw_vp_fv_sub |> 
  filter(`Development Stage` == "Stage 3")

gfw_afe3_id_sub <- gfw_afe_id_sub |> 
  filter(`Development Stage` == "Stage 3")
```

## more sniff tests - visualization checks

```{r}
# histograms

gfw_afe_id_sub |> 
  ggplot(aes(x=`Apparent Fishing Hours`)) + geom_histogram(binwidth = 0.5) + theme_bw()

gfw_vp_fv_sub |> 
  ggplot(aes(x=`Vessel Presence Hours`)) + geom_histogram(binwidth = 1) + theme_bw()

# how to handle one-inflation in data sets?
```

```{r}
# check dist of records by month

gfw_vp_fv_sub |> 
  ggplot(aes(x = `Year Month`)) +
  geom_bar(fill = "steelblue") +
  theme_bw()

gfw_afe_id_sub |> 
  ggplot(aes(x = `Year Month`)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # why sudden jump in mid/late 2024 and 2025?

# check by dev stage

gfw_vp_fv_sub |> 
  ggplot(aes(x = `Development Stage`)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # steady increase a result of just more general fishing activity over time?

gfw_afe_id_sub |> 
  ggplot(aes(x = `Development Stage`)) +
  geom_bar(fill = "steelblue") +
  theme_bw() # BIG jump in dev stage 3
```

```{r}
# check dist of unique vessel IDs by month

IDs_perMonth <- gfw_vp_fv_sub |> 
  group_by(`Year Month`) |> 
  summarise(unique_id_count = n_distinct(`Vessel ID`))

IDs_perMonth |> 
  ggplot(aes(x = as.factor(`Year Month`), y = unique_id_count)) +
  geom_col(fill = "steelblue") +
  theme_bw() # trend seems consistent over time... not seeing any big spikes in later years that should indicate more vessels fishing in the area

# check by dev stage

IDs_perDev <- gfw_vp_fv_sub |> 
  group_by(`Development Stage`) |> 
  summarise(unique_id_count = n_distinct(`Vessel ID`))

IDs_perDev |> 
  ggplot(aes(x = as.factor(`Development Stage`), y = unique_id_count)) +
  geom_col(fill = "steelblue") +
  theme_bw() # slight decrease in no. of vessels from stage 1 to 2 then small rebound in 3
```

## Export data sets

```{r}
# all vp

write_csv(gfw_vp_id, "data/external/gfw/gfw_vp_id.csv")

# orsted vessels 

write_csv(gfw_vp_orstedVessels, "data/processed/gfw/gfw_vp_orstedVessels.csv")

# vp_fv

write_csv(gfw_vp_fv, "data/processed/gfw/gfw_vp_fv.csv")
write_csv(gfw_vp_fv_sub, "data/processed/gfw/gfw_vp_fv_sub.csv")
write_csv(gfw_vp1_fv_sub, "data/processed/gfw/gfw_vp1_fv_sub.csv")
write_csv(gfw_vp2_fv_sub, "data/processed/gfw/gfw_vp2_fv_sub.csv")
write_csv(gfw_vp3_fv_sub, "data/processed/gfw/gfw_vp3_fv_sub.csv")

# afe_id

write_csv(gfw_afe_id, "data/external/gfw/gfw_afe_id.csv")
write_csv(gfw_afe_id_sub, "data/processed/gfw/gfw_afe_id_sub.csv")
write_csv(gfw_afe1_id_sub, "data/processed/gfw/gfw_afe1_id_sub.csv")
write_csv(gfw_afe2_id_sub, "data/processed/gfw/gfw_afe2_id_sub.csv")
write_csv(gfw_afe3_id_sub, "data/processed/gfw/gfw_afe3_id_sub.csv")
```