---
title: "MM GFW Vessel Classification: False Positives and False Negatives"
author: "Malachy McCaffrey"
date: today
format: html
params:
  run_gfw_api: false  # true to look up uncached candidates in the GFW identity API
  # Path overrides (see r/R/paths.R); leave blank to use the project paths.
  decisions: ""
  gear: ""
  crosswalk: ""
  review: ""
  evidence: ""
  processed_gfw: ""
  interim_gfw: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# run from the repo root (the folder holding renv.lock) so the relative paths
# in r/R/paths.R resolve whether the qmd is rendered or run chunk by chunk
root <- getwd()
while (!file.exists(file.path(root, "renv.lock")) && dirname(root) != root) root <- dirname(root)
knitr::opts_knit$set(root.dir = root)
```

## Objective

Identify and account for vessels that GFW misclassifies in the raw VP and AFE
pulls, following the hybrid approach in `r/scratchpad/vessel_filtering_plan.qmd`:

- a **decisions file** (`references/vessel_decisions.csv`, tracked in git) is
  the only file edited by hand; together with `references/vessel_removals.csv`
  it generates the **MMSI crosswalk** that classifies every record
- a **screen** of the raw data produces review lists
- the **GFW vessel-identity API** and **NOAA GARFO permit lists** supply evidence
- **three analysis sets** (Strict / Primary / Inclusive) carry what stays
  uncertain into the results

Two separate problems are handled:

- **Hull identity** -- is the vessel a fishing vessel at all? GFW types some
  fishing vessels OTHER, GEAR or PASSENGER (false negatives) and some
  non-fishing vessels FISHING (false positives).
- **Activity** -- is a fishing vessel fishing at this moment? Charter guard and
  survey work by fishing hulls is classed as fishing by GFW (false positives).
  This depends on dates, so crosswalk rows carry a validity window.

## How to use this workflow

Every run rebuilds everything from tracked inputs, so the outputs depend only
on the raw pull, the GARFO files, the API cache and these files:

| File | Role |
|---|---|
| `references/vessel_removals.csv` | Ørsted charter and other known removals (input) |
| `references/vessel_decisions.csv` | Your hull and activity decisions (input, edited by hand) |
| `references/vessel_gear.csv` | Your gear corrections (input, edited by hand) |
| `references/garfo_permit_codes.csv` | GARFO permit code meanings (input) |
| `references/vessel_crosswalk.csv` | Generated each run from the removals and decisions -- do not edit |
| `data/processed/vessel_review/vessel_review.csv` | Generated each run: hull/activity candidates, evidence and a suggested decision |
| `data/processed/vessel_review/gear_review.csv` | Generated each run: fishing vessels whose gear needs checking, with a suggested gear class |

Each pass:

1. **Run with the API off** (the default). Check the list sizes and screening
   thresholds.
2. **Run once with the API on**: render with `-P run_gfw_api:true`, or set
   `run_gfw_api: true` in the header. Lookups are cached in
   `data/raw/vessel_evidence/`, so each Vessel ID is queried once; set it back
   to `false` afterwards.
3. **Review** the two sheets. Their first columns are a suggestion in the same
   layout as the matching input file (`vessel_review.csv` →
   `vessel_decisions.csv`; `gear_review.csv` → `vessel_gear.csv`). For each row
   you decide on, correct the suggestion if needed and set `decision` to
   `accept` or `reject`. Leave `decision` blank to skip a row.
4. **Re-run.** The run first copies every row with a `decision` from the two
   sheets into their input files, then regenerates the crosswalk and both
   sheets, so decided vessels drop off. You can also edit the input files
   directly.
5. **Commit `vessel_decisions.csv` and `vessel_gear.csv`** after each review
   session; they are the record of every decision.

Hull and activity decisions are kept separate from gear corrections, so a
vessel's gear can be fixed without touching its charter exclusion (e.g. a
removal-list charter vessel whose GFW gear is wrong).

If you open the review sheet in Excel, import it with every column as Text
(Data → From Text/CSV) and save as CSV UTF-8. Opened directly, Excel rewrites
dates (e.g. `2024-04-26` → `4/26/2024`) and long IDs, and the crosswalk
validation then stops the run.

## Libraries and shared code

```{r}
library(tidyverse)
library(sf)
library(gfwr)

source("r/R/paths.R")
source("r/R/gfw_helpers.R")

# Quarto parameters (header); `params` is read-only while rendering, and does
# not exist when chunks are run interactively
qparams     <- if (exists("params")) params else list()
paths       <- gfw_paths_with(qparams)
run_gfw_api <- isTRUE(qparams$run_gfw_api)
```

## Parameters

```{r}
# Date of the raw pull, recorded by gfw_vp_afe_datapull.qmd in pull_info.csv.
# Pulls made before that file existed fall back to the raw VP file's
# modification date, with a warning.
pull_date <- if (file.exists(paths$pull_info)) {
  as.Date(read_csv(paths$pull_info, col_types = cols(.default = col_character()))$pull_date[1])
} else {
  warning(paths$pull_info, " not found; using the modification date of ", paths$raw_vp,
          " as the pull date", call. = FALSE)
  # local date; as.Date() on a time would use UTC and can shift the day
  as.Date(format(file.mtime(paths$raw_vp), "%Y-%m-%d"))
}
pull_date

# screening thresholds -- record any change in the methods
screen <- list(
  min_s3_hours     = 50,   # charter screen: minimum Stage 3 VP hours
  min_lease_share  = 0.25, # charter screen: share of hours inside SFW/RWF/SRW
  max_prior_hours  = 0,    # charter screen: Stage 1-2 hours for a "no history" burst
  min_lease_shift  = 0.25, # charter screen: rise in lease share, Stage 1-2 to 3
  fp_max_days      = 30,   # FP screen: short foreign-flag visit, active days
  fp_min_h_day     = 12,   # FP screen: hours per active day
  length_tolerance = 0.15, # GARFO vs GFW registry length agreement (share)
  oa_bundle_min    = 5     # open-access commercial permits that mark a party-boat bundle
)

api_batch_size <- 20 # Vessel IDs per API request
```

## Read raw data

```{r}
# read_gfw() (r/R/gfw_helpers.R) types the hours, derives Year Month and
# Development Stage, and snaps Lat/Lon to the 0.01 degree cell_id
vp  <- read_gfw(paths$raw_vp)
afe <- read_gfw(paths$raw_afe)

stopifnot("records with no development stage" =
            !anyNA(vp$`Development Stage`) && !anyNA(afe$`Development Stage`))

total_vp  <- sum(vp$hours)
total_afe <- sum(afe$hours)
```

```{r}
# GFW labels per MMSI as of this pull (most frequent, in case of a split)
gfw_labels <- bind_rows(vp, afe) |>
  filter(!is.na(MMSI)) |>
  count(MMSI, `Vessel Type`, `Gear Type`) |>
  group_by(MMSI) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(MMSI, gfw_vessel_type = `Vessel Type`, gfw_gear_type = `Gear Type`)
```

```{r}
# lease flag per grid cell (cell centre inside SFW, RWF or SRW)
grid <- st_read(paths$grid, quiet = TRUE)

leases <- paths$leases |>
  map(\(p) st_read(p, quiet = TRUE) |> st_transform(st_crs(grid)) |> st_union()) |>
  reduce(st_union)

cell_lease <- tibble(
  cell_id  = grid$cell_id,
  in_lease = lengths(st_intersects(st_centroid(st_geometry(grid)), leases)) > 0)

stopifnot("records outside the fishnet grid" =
            all(c(vp$cell_id, afe$cell_id) %in% cell_lease$cell_id))
```

## Phase 1: Decisions and the crosswalk

The crosswalk has one row per MMSI per validity period. A record is matched to
the row whose `valid_from`–`valid_to` window contains its date (blank =
open-ended). Records with no matching row keep GFW's label. Windows for one
MMSI must not overlap, so a fishing hull GFW missed that also took a charter
needs two rows: one before the charter (`activity_excluded = FALSE`) and one
for the charter (`TRUE`).

| Column | Values |
|---|---|
| `MMSI` | 9 digits (any format for `gear_marker`) |
| `valid_from`, `valid_to` | ISO dates (`YYYY-MM-DD`); blank = open-ended |
| `hull_class` | `fishing` / `non_fishing` / `gear_marker` / `unknown` |
| `activity_excluded` | `TRUE` = records in the window are not fishing (charter work) |
| `Confidence` | `confirmed` / `probable` / `possible` |
| `Identified_By`, `evidence_ref`, `Notes` | Where the decision came from |
| `gfw_vessel_type`, `gfw_gear_type`, `pull_date` | GFW's labels when the decision was made |

`vessel_decisions.csv` has the same columns plus `decision` (`accept` /
`reject`) and `decided_on`. The crosswalk is built as:

- **Removal-list rows.** Charter vessels (`Scope == "stages_2_3"`) become
  fishing hulls with an activity exclusion from the start of Stage 2;
  `all_stages` vessels become non-fishing hulls.
- **Accepted decisions.** Any decision for an MMSI on the removal list
  replaces that vessel's removal-list rows -- this is how to give a charter
  vessel its real charter dates. A `reject` for a removal-list MMSI removes
  it from the crosswalk.
- **Rejected decisions** add no row: GFW's label stands, and the vessel no
  longer appears in the review sheet.

Gear is not part of the crosswalk. It is corrected separately in
`references/vessel_gear.csv` (see *Gear corrections* below), so fixing a
vessel's gear never touches its hull or charter treatment.

```{r}
cw_cols <- c("MMSI", "valid_from", "valid_to", "Vessel_Name", "hull_class",
             "activity_excluded", "Operation", "Identified_By",
             "Confidence", "evidence_ref", "gfw_vessel_type", "gfw_gear_type",
             "pull_date", "Notes")
decision_cols <- c("decision", "decided_on", cw_cols)

gear_cols <- c("decision", "decided_on", "MMSI", "valid_from", "valid_to",
               "Vessel_Name", "gear_class", "basis", "evidence_ref",
               "gfw_gear_type", "pull_date", "Notes")

read_chr <- function(path) {
  read_csv(path, col_types = cols(.default = col_character()), na = c("", "NA"))
}

empty_tbl <- function(cols) as_tibble(set_names(rep(list(character()), length(cols)), cols))

# Copies rows marked accept/reject in a generated review sheet into its input
# file, skipping MMSIs the input file already has, and returns the input file.
collect_marked <- function(sheet, input, cols) {
  current <- if (file.exists(input)) read_chr(input) else empty_tbl(cols)
  stopifnot("input file is missing columns" = all(cols %in% names(current)))

  marked <- if (file.exists(sheet)) {
    read_chr(sheet) |>
      filter(decision %in% c("accept", "reject")) |>
      mutate(decided_on = coalesce(decided_on, as.character(Sys.Date()))) |>
      select(all_of(cols))
  } else {
    current[0, ]
  }

  skipped <- semi_join(marked, current, by = "MMSI")
  if (nrow(skipped) > 0) {
    warning(nrow(skipped), " rows in ", sheet, " were skipped because their MMSI ",
            "is already in ", input, "; edit that file to change them", call. = FALSE)
  }

  new_rows <- anti_join(marked, current, by = "MMSI")
  if (nrow(new_rows) > 0) {
    current <- bind_rows(current, new_rows) |> select(all_of(cols))
    write_csv(current, input, na = "")
    message("Added ", nrow(new_rows), " rows to ", input)
  }

  stopifnot("decision must be accept or reject" =
              all(current$decision %in% c("accept", "reject")))
  current
}
```

### Collect decisions from the last review sheets

```{r}
review_path      <- file.path(paths$review, "vessel_review.csv")
gear_review_path <- file.path(paths$review, "gear_review.csv")

decisions   <- collect_marked(review_path, paths$decisions, decision_cols)
gear_inputs <- collect_marked(gear_review_path, paths$gear, gear_cols)

count(decisions, decision)
count(gear_inputs, decision)
```

### Build and validate the crosswalk

```{r}
lo_date <- as.Date("1900-01-01") # stand-ins for open-ended windows
hi_date <- as.Date("2100-12-31")

validate_crosswalk <- function(cw) {
  overlaps <- cw |>
    mutate(f = coalesce(valid_from, lo_date), t = coalesce(valid_to, hi_date)) |>
    arrange(MMSI, f) |>
    group_by(MMSI) |>
    filter(lead(f) <= t | lag(t) >= f) |>
    pull(MMSI) |>
    unique()

  problems <- c(
    if (any(!cw$hull_class %in% c("fishing", "non_fishing", "gear_marker", "unknown")))
      "unknown hull_class",
    if (any(!cw$Confidence %in% c("confirmed", "probable", "possible")))
      "unknown Confidence",
    if (anyNA(cw$activity_excluded))
      "activity_excluded must be TRUE or FALSE",
    if (any(cw$hull_class != "gear_marker" & !grepl("^[0-9]{9}$", cw$MMSI)))
      "MMSI is not 9 digits (allowed only for gear_marker)",
    if (any(cw$valid_from > cw$valid_to, na.rm = TRUE))
      "valid_from after valid_to",
    if (length(overlaps) > 0)
      paste0("overlapping windows for MMSI: ", paste(overlaps, collapse = ", "))
  )

  if (length(problems) > 0) {
    stop("Invalid crosswalk (check ", paths$decisions, "):\n  - ",
         paste(problems, collapse = "\n  - "), call. = FALSE)
  }
  invisible(cw)
}

stage2_start <- as.character(dev_stages$start[dev_stages$stage == "Stage 2"])

removal_rows <- read_chr(paths$removals) |>
  filter(!MMSI %in% decisions$MMSI) |>
  transmute(MMSI, Vessel_Name,
            valid_from = if_else(Scope == "stages_2_3", stage2_start, NA_character_),
            valid_to   = NA_character_,
            hull_class = if_else(Scope == "stages_2_3", "fishing", "non_fishing"),
            activity_excluded = if_else(Scope == "stages_2_3", "TRUE", "FALSE"),
            Operation, Identified_By, Confidence,
            evidence_ref = NA_character_,
            # the pull the crosswalk was built against (pull_info.csv)
            pull_date = as.character(.env$pull_date),
            Notes) |>
  # GFW's labels for these vessels in that pull
  left_join(gfw_labels, by = "MMSI")

cw_out <- bind_rows(removal_rows,
                    decisions |> filter(decision == "accept") |> select(all_of(cw_cols))) |>
  select(all_of(cw_cols)) |>
  arrange(MMSI, valid_from)

cw <- cw_out |>
  mutate(valid_from = as.Date(valid_from),
         valid_to   = as.Date(valid_to),
         activity_excluded = as.logical(activity_excluded)) |>
  validate_crosswalk()

write_csv(cw_out, paths$crosswalk, na = "")

count(cw, hull_class, activity_excluded, Confidence)
```

### Gear corrections

`references/vessel_gear.csv` overrides GFW's gear for the records of a vessel
inside each row's `valid_from`–`valid_to` window (blank = open-ended). An
`accept` row sets `gear_class` (`MOBILE` / `FIXED` / `POLE_AND_LINE` /
`UNRESOLVED`); a `reject` row records that GFW's gear was checked and kept, so
the vessel leaves the gear review sheet. `basis` says where the gear came from
(e.g. `analyst knowledge`, `GARFO permit`, `GFW registry`). Every record's
gear class is then the accepted correction if there is one, otherwise GFW's
`Gear Type` through the gear map in `r/R/gfw_helpers.R`.

```{r}
gear <- gear_inputs |>
  filter(decision == "accept") |>
  mutate(valid_from = as.Date(valid_from), valid_to = as.Date(valid_to))

gear_overlaps <- gear |>
  mutate(f = coalesce(valid_from, lo_date), t = coalesce(valid_to, hi_date)) |>
  arrange(MMSI, f) |>
  group_by(MMSI) |>
  filter(lead(f) <= t | lag(t) >= f) |>
  pull(MMSI) |>
  unique()

gear_problems <- c(
  if (any(!gear$gear_class %in% c("MOBILE", "FIXED", "POLE_AND_LINE", "UNRESOLVED")))
    "gear_class must be MOBILE, FIXED, POLE_AND_LINE or UNRESOLVED",
  if (any(gear$valid_from > gear$valid_to, na.rm = TRUE))
    "valid_from after valid_to",
  if (length(gear_overlaps) > 0)
    paste0("overlapping windows for MMSI: ", paste(gear_overlaps, collapse = ", "))
)
if (length(gear_problems) > 0) {
  stop("Invalid ", paths$gear, ":\n  - ", paste(gear_problems, collapse = "\n  - "),
       call. = FALSE)
}

count(gear, gear_class, basis)
```

## Phase 2: Screen the raw data

### Per-MMSI profile

```{r}
profile_mmsi <- function(vp, afe, cell_lease) {
  v <- vp |> filter(!is.na(MMSI)) |> left_join(cell_lease, by = "cell_id")
  a <- afe |> filter(!is.na(MMSI))

  stage_hours <- function(h, stage, s) sum(h[stage == s])

  pv <- v |>
    group_by(MMSI) |>
    summarise(
      names     = paste(sort(unique(na.omit(`Vessel Name`))), collapse = " / "),
      flag      = first(na.omit(Flag)),
      call_sign = first(na.omit(CallSign)),
      vp_hours  = sum(hours),
      s1 = stage_hours(hours, `Development Stage`, "Stage 1"),
      s2 = stage_hours(hours, `Development Stage`, "Stage 2"),
      s3 = stage_hours(hours, `Development Stage`, "Stage 3"),
      days  = n_distinct(date),
      h_day = vp_hours / days,
      lease_share     = sum(hours[in_lease]) / vp_hours,
      lease_share_s12 = if (s1 + s2 > 0)
        sum(hours[in_lease & `Development Stage` != "Stage 3"]) / (s1 + s2) else NA_real_,
      lease_share_s3  = if (s3 > 0)
        sum(hours[in_lease & `Development Stage` == "Stage 3"]) / s3 else NA_real_,
      s3_lease_first = suppressWarnings(min(date[in_lease & `Development Stage` == "Stage 3"])),
      s3_lease_last  = suppressWarnings(max(date[in_lease & `Development Stage` == "Stage 3"])),
      first = min(date),
      last  = max(date),
      .groups = "drop") |>
    mutate(across(c(s3_lease_first, s3_lease_last), \(x) if_else(is.finite(x), x, NA)))

  pa <- a |>
    group_by(MMSI) |>
    summarise(afe_hours = sum(hours),
              afe_s3 = sum(hours[`Development Stage` == "Stage 3"]),
              .groups = "drop")

  pv |>
    left_join(pa, by = "MMSI") |>
    left_join(gfw_labels, by = "MMSI") |>
    mutate(afe_hours = coalesce(afe_hours, 0),
           afe_s3    = coalesce(afe_s3, 0),
           mmsi_valid = grepl("^[0-9]{9}$", MMSI),
           battery_call_sign = str_detect(coalesce(call_sign, ""), "%"))
}

profiles <- profile_mmsi(vp, afe, cell_lease)

nrow(profiles)
```

### GARFO permit matches

The GARFO permit lists (`data/raw/vessel_evidence/registries/`, one file per
permit year plus the HMS/tuna file) identify vessels only by name, permit
number and hull ID. Neither permit number nor hull ID appears in AIS, so the
only link is the vessel name. Names collide (many GARFO names belong to more
than one hull, and placeholders such as UNNAMED and SKIFF are common), so a
name match is only a lead:

| `garfo_match` | Meaning |
|---|---|
| `unique_years` | One GARFO hull and one US-flag MMSI share the name, and a permit year falls within the MMSI's AIS activity |
| `unique_no_year_overlap` | Unique, but no permit year overlaps the AIS activity |
| `ambiguous` | The name belongs to several hulls, or to several MMSIs active at the same time. MMSIs sharing a name whose activity never overlaps (a changed MMSI, or a replacement vessel with the same name) are not counted as ambiguous (`garfo_mmsi_sequence`); the permit years must still overlap the MMSI's activity |
| `none` | No match |

Phase 4 adds a length check once GFW registry lengths are available. A name
match is never suggested as `confirmed`. To confirm one, look up the hull ID in
the USCG PSIX search on CGMIX (numeric hull IDs are USCG official numbers) and
check that the vessel's call sign matches the AIS call sign, then record the
hull ID in `evidence_ref`. State-registered hull IDs (e.g. `MS6930PP`) cannot
be linked to a call sign this way.

Only vessel fields are read; owner names, telephone numbers and addresses are
never loaded.

```{r}
garfo_personal <- c("OWNER NAME", "TELEPHONE", "ADDRESS LINE 1", "ADDRESS LINE 2",
                    "CITY", "STATE", "ZIP CODE", "HOME PORT CITY",
                    "PRINCIPAL PORT CITY")

garfo_vessel_cols <- c("PERMIT", "HMS PERMIT", "PERMIT YEAR", "VESSEL NAME",
                       "HULL ID", "HOME PORT STATE", "PRINCIPAL PORT STATE",
                       "LENGTH", "GROSS TONS", "HORSEPOWER", "MAX TRAP LIMIT",
                       "CATEGORY", "CML CHB ENDORSEMENT", "ISSUEDDATE")

placeholder_names <- c("UNNAMED", "NONAME", "NONE", "NA", "SKIFF", "UNKNOWN")

norm_name <- function(x) {
  x |>
    toupper() |>
    str_replace_all("&", " AND ") |>
    str_remove("^[[:space:]]*(F[[:space:]]*/[[:space:]]*V|F[.]V[.]|FV)[[:space:]]+") |>
    str_remove_all("[^A-Z0-9]")
}

col_or_na <- function(d, col) if (col %in% names(d)) d[[col]] else rep(NA_character_, nrow(d))

read_garfo <- function(f) {
  d <- read_csv(f, col_types = cols(.default = col_character()),
                col_select = !any_of(garfo_personal),
                na = c("", " ", "NA", "None"), show_col_types = FALSE)
  names(d) <- str_remove(names(d), paste0("^", intToUtf8(0xFEFF))) # byte-order mark

  hms <- "HMS PERMIT" %in% names(d)

  permits <- tibble(
    file        = basename(f),
    source      = if (hms) "garfo_hms" else "garfo_permits",
    permit      = coalesce(col_or_na(d, "PERMIT"), col_or_na(d, "HMS PERMIT")),
    permit_year = suppressWarnings(as.integer(col_or_na(d, "PERMIT YEAR"))),
    vessel_name = col_or_na(d, "VESSEL NAME"),
    hull_id     = col_or_na(d, "HULL ID"),
    home_port_state = col_or_na(d, "HOME PORT STATE"),
    length_ft   = suppressWarnings(as.numeric(col_or_na(d, "LENGTH"))),
    trap_limit  = suppressWarnings(as.numeric(col_or_na(d, "MAX TRAP LIMIT"))))

  # one row per permit-year x fishery x code. Annual files hold one column
  # per fishery with comma-separated codes; the HMS file has a CATEGORY.
  fishery <- setdiff(names(d), garfo_vessel_cols)
  codes <- if (hms) {
    # a charter/headboat permit endorsed for commercial sale is a second,
    # commercial code for the same permit
    bind_rows(
      tibble(permit = permits$permit, permit_year = permits$permit_year,
             fishery = "HMS", code = col_or_na(d, "CATEGORY")),
      tibble(permit = permits$permit, permit_year = permits$permit_year,
             fishery = "HMS",
             code = if_else(col_or_na(d, "CML CHB ENDORSEMENT") %in% "Y",
                            "CHB COMMERCIAL SALE ENDORSEMENT", NA_character_)))
  } else {
    bind_cols(select(permits, permit, permit_year), d[fishery]) |>
      pivot_longer(all_of(fishery), names_to = "fishery", values_to = "code",
                   values_drop_na = TRUE) |>
      separate_longer_delim(code, ",")
  }
  codes <- codes |>
    mutate(file = basename(f), code = str_trim(code)) |>
    filter(!is.na(code), code != "")

  list(permits = permits, codes = codes)
}

garfo_files <- list.files(paths$registries,
                          pattern = "^garfo.*[.]csv$", full.names = TRUE)

garfo_read  <- map(garfo_files, read_garfo)
garfo_all   <- map(garfo_read, "permits") |> list_rbind()
garfo_codes <- map(garfo_read, "codes") |> list_rbind()

# rows broken in the source file (e.g. shifted fields) have no valid year
malformed <- with(garfo_all, is.na(permit_year) | !between(permit_year, 2000, 2100) |
                    is.na(hull_id) | is.na(vessel_name))
message("GARFO rows read: ", nrow(garfo_all), "; malformed rows dropped: ", sum(malformed))

garfo_all   <- garfo_all[!malformed, ]
garfo_codes <- semi_join(garfo_codes, garfo_all, by = c("file", "permit", "permit_year"))
```

### GARFO permit codes

Permit codes are fishery-specific. The meanings live in
`references/garfo_permit_codes.csv`, one row per fishery and code, with
`permit_type` (`commercial` / `for_hire` / `recreational` / `unknown`), an
optional `gear_hint` (`FIXED` / `MOBILE`), a `description`, and the `basis`
and `source` for each mapping (NOAA Fisheries permit pages). Category 2 is
charter/party in every fishery that has one (bluefish, black sea bass, scup,
summer flounder, tilefish, lobster, squid/mackerel/butterfish), and tilefish 3
is the private recreational permit; otherwise codes are fishery-specific (e.g.
squid/mackerel/butterfish 3 = incidental catch, 4 = Atlantic mackerel open
access). Codes missing from the table are listed below and treated as
`unknown`.

For commercial permits, `open_access` marks those anyone can obtain, and
`oa_directed` separates open-access permits for a directed fishery (e.g.
spiny dogfish, bluefish, tilefish, Handgear B, HMS General) from incidental
ones (monkfish E, squid/butterfish incidental, multispecies J). The HMS file's
`CML CHB ENDORSEMENT = Y` (a charter/headboat permit endorsed for commercial
sale) is read as an extra open-access commercial code.

```{r}
permit_codes <- read_csv(paths$permit_codes, col_types = cols(.default = col_character()),
                         na = "")

unmapped <- garfo_codes |>
  count(fishery, code, name = "permit_years") |>
  anti_join(permit_codes, by = c("fishery", "code"))

if (nrow(unmapped) > 0) {
  warning(nrow(unmapped), " GARFO codes are missing from ", paths$permit_codes,
          " and are treated as unknown", call. = FALSE)
}
unmapped

# permit type and gear hints per permit-year
permit_flags <- garfo_codes |>
  left_join(permit_codes, by = c("fishery", "code")) |>
  mutate(fc = paste0(fishery, ":", code),
         comm_oa = permit_type %in% "commercial" & open_access %in% "TRUE") |>
  group_by(file, permit, permit_year) |>
  summarise(
    # commercial permits that are not open access (limited access, moratorium)
    commercial_la = any(permit_type %in% "commercial" & !comm_oa),
    commercial_oa = any(comm_oa),
    for_hire      = any(permit_type %in% "for_hire"),
    recreational  = any(permit_type %in% "recreational"),
    fixed_hint    = any(gear_hint %in% "FIXED"),
    mobile_hint   = any(gear_hint %in% "MOBILE"),
    oa_permits    = paste(fc[comm_oa], collapse = "|"),
    oa_directed_permits = paste(fc[comm_oa & oa_directed %in% "TRUE"], collapse = "|"),
    categories    = paste(fc, collapse = "; "),
    .groups = "drop")

garfo <- garfo_all |>
  left_join(permit_flags, by = c("file", "permit", "permit_year")) |>
  mutate(across(c(commercial_la, commercial_oa, for_hire, recreational, fixed_hint, mobile_hint),
                \(x) coalesce(x, FALSE)),
         fixed_hint   = fixed_hint | coalesce(trap_limit > 0, FALSE),
         nkey         = norm_name(vessel_name),
         hull_is_uscg = str_detect(hull_id, "^[0-9]+$")) |>
  filter(!nkey %in% placeholder_names, nchar(nkey) >= 3)

garfo |>
  count(source, commercial_la, commercial_oa, for_hire, recreational)
```

```{r}
# AIS names per MMSI; GARFO vessels are US, so match US- or unflagged MMSIs
ais_names <- bind_rows(vp, afe) |>
  filter(!is.na(MMSI), !is.na(`Vessel Name`), Flag %in% c("USA", NA)) |>
  distinct(MMSI, `Vessel Name`) |>
  mutate(nkey = norm_name(`Vessel Name`)) |>
  filter(!nkey %in% placeholder_names, nchar(nkey) >= 3) |>
  distinct(MMSI, nkey)

# MMSIs sharing a name whose AIS activity never overlaps are one vessel that
# changed MMSI, or a replacement vessel given the same name (e.g. CAILYN &
# MAREN 2021-2025, then CAILYN AND MAREN II from late 2025). Either way they
# do not make the name ambiguous; the permit-year check still has to tie the
# GARFO hull to the right MMSI. An MMSI with no profile counts as overlapping.
ais_name_counts <- ais_names |>
  left_join(select(profiles, MMSI, first, last), by = "MMSI") |>
  arrange(nkey, first) |>
  group_by(nkey) |>
  summarise(mmsi_with_name = n_distinct(MMSI),
            mmsi_overlap   = anyNA(first) ||
              any(lead(as.numeric(first)) <= cummax(as.numeric(last)), na.rm = TRUE),
            .groups = "drop")

name_counts <- full_join(
  garfo |> group_by(nkey) |> summarise(hulls_with_name = n_distinct(hull_id)),
  ais_name_counts,
  by = "nkey")

garfo_by_mmsi <- ais_names |>
  inner_join(garfo, by = "nkey", relationship = "many-to-many") |>
  left_join(name_counts, by = "nkey") |>
  left_join(select(profiles, MMSI, first, last), by = "MMSI") |>
  group_by(MMSI) |>
  summarise(
    garfo_hulls = n_distinct(hull_id),
    garfo_max_hulls_with_name = max(hulls_with_name),
    garfo_max_mmsi_with_name  = max(mmsi_with_name),
    garfo_mmsi_overlap  = any(mmsi_with_name > 1 & mmsi_overlap),
    garfo_mmsi_sequence = any(mmsi_with_name > 1 & !mmsi_overlap),
    year_overlap     = any(permit_year >= year(first) & permit_year <= year(last)),
    garfo_years      = paste(sort(unique(permit_year)), collapse = ","),
    garfo_hull_id    = paste(sort(unique(hull_id)), collapse = "|"),
    hull_is_uscg     = all(hull_is_uscg),
    garfo_permits    = paste(sort(unique(permit)), collapse = "|"),
    garfo_sources    = paste(sort(unique(source)), collapse = "|"),
    garfo_length_ft  = median(length_ft, na.rm = TRUE),
    garfo_home_port  = paste(sort(unique(na.omit(home_port_state))), collapse = "|"),
    garfo_categories = str_trunc(paste(unique(na.omit(categories)), collapse = " / "), 300),
    any_commercial_la = any(commercial_la),
    any_commercial_oa = any(commercial_oa),
    any_for_hire      = any(for_hire),
    any_recreational  = any(recreational),
    any_fixed         = any(fixed_hint),
    any_mobile        = any(mobile_hint),
    garfo_oa_permits  = paste(sort(unique(unlist(str_split(oa_permits, fixed("|"))))) |>
                               setdiff(""), collapse = "|"),
    garfo_oa_directed = paste(sort(unique(unlist(str_split(oa_directed_permits, fixed("|"))))) |>
                               setdiff(""), collapse = "|"),
    .groups = "drop") |>
  mutate(garfo_length_ft = if_else(is.nan(garfo_length_ft), NA_real_, garfo_length_ft),
         garfo_match = case_when(
           garfo_hulls > 1 | garfo_mmsi_overlap ~ "ambiguous",
           !year_overlap                        ~ "unique_no_year_overlap",
           TRUE                                 ~ "unique_years"),
         garfo_n_oa = if_else(garfo_oa_permits == "", 0L,
                              str_count(garfo_oa_permits, fixed("|")) + 1L),
         # Open-access commercial permits are easy to obtain, so for-hire
         # vessels often hold them without fishing commercially. Keep that
         # combination as its own type rather than folding it into either
         # "for-hire" or "commercial + for-hire".
         garfo_permit_type = case_when(
           any_commercial_la & any_for_hire ~ "commercial + for-hire",
           any_commercial_oa & any_for_hire ~ "for-hire + open-access commercial",
           any_for_hire                     ~ "for-hire",
           any_commercial_la | any_commercial_oa ~ "commercial",
           any_recreational                 ~ "recreational only",
           TRUE                             ~ "unmapped codes only"),
         garfo_oa_bundle = garfo_permit_type == "for-hire + open-access commercial" &
                           garfo_n_oa >= screen$oa_bundle_min,
         # gear hint only when it is unambiguous; a for-hire vessel with a
         # small set of open-access commercial permits may fish commercially
         garfo_gear_hint = case_when(
           garfo_permit_type == "for-hire"           ~ "POLE_AND_LINE",
           garfo_oa_bundle                           ~ "POLE_AND_LINE",
           any_for_hire                              ~ NA_character_,
           any_fixed & !any_mobile                   ~ "FIXED",
           any_mobile & !any_fixed                   ~ "MOBILE",
           TRUE                                      ~ NA_character_)) |>
  select(-starts_with("any_"))

profiles <- profiles |>
  left_join(garfo_by_mmsi, by = "MMSI") |>
  mutate(garfo_match = coalesce(garfo_match, "none"))
```

How often does a name match succeed for vessels whose fishing status is
already known? US-flag vessels GFW types FISHING, and the fishing hulls in the
crosswalk, set the expectation; matches among non-FISHING vessels are the
leads.

```{r}
profiles |>
  filter(flag %in% "USA") |>
  mutate(group = case_when(
    MMSI %in% cw$MMSI[cw$hull_class == "fishing"] ~ "crosswalk fishing hull",
    gfw_vessel_type %in% "FISHING"                ~ "GFW FISHING",
    TRUE                                          ~ "GFW non-FISHING")) |>
  count(group, garfo_match) |>
  group_by(group) |>
  mutate(share = round(n / sum(n), 3)) |>
  ungroup()
```

### Candidate lists

Each list excludes vessels already in the crosswalk or in the decisions file.

1. **Missed fishing vessels** -- typed other than FISHING, and either US-flag
   with a call sign or with a unique, year-overlapping GARFO name match. This
   whole pool is looked up in the GFW API, but only vessels with evidence
   reach the review sheet (Phase 4).
2. **Charter activity** -- typed FISHING with a lease-concentrated Stage 3:
   either no Stage 1–2 history (a burst), or a lease share that rose from
   Stages 1–2 to Stage 3 (catches established vessels that took a charter, the
   *Virginia Wave* blind spot).
3. **Non-fishing vessels typed FISHING** -- research or survey names, short
   intense foreign-flag visits, or foreign-flag `INCONCLUSIVE` gear.
4. **Gear markers and malformed MMSIs** -- call signs holding battery levels
   (e.g. ONWA33NT, likely AIS net buoys), or MMSIs that are not 9 digits
   (some are vessels with misconfigured transponders, e.g. ROYAL with call sign
   WDD3653, so these are suggested as `unknown` for review).

```{r}
charter_screen <- function(p) {
  p |>
    mutate(burst = s1 + s2 <= screen$max_prior_hours &
                   lease_share >= screen$min_lease_share,
           shift = !is.na(lease_share_s12) & !is.na(lease_share_s3) &
                   lease_share_s3 - lease_share_s12 >= screen$min_lease_shift,
           charter_flag = s3 >= screen$min_s3_hours & (burst | shift))
}

open_profiles <- profiles |> filter(!MMSI %in% c(cw$MMSI, decisions$MMSI))

cand_missed <- open_profiles |>
  filter(!gfw_vessel_type %in% "FISHING", mmsi_valid,
         (flag %in% "USA" & !is.na(call_sign)) | garfo_match == "unique_years") |>
  mutate(list = "1_missed_fishing",
         reason = if_else(garfo_match == "unique_years",
                          "typed non-FISHING; unique GARFO name match",
                          "typed non-FISHING; US flag with call sign"))

cand_charter <- charter_screen(open_profiles) |>
  filter(gfw_vessel_type %in% "FISHING", charter_flag) |>
  mutate(list = "2_charter_activity",
         reason = case_when(
           burst & shift ~ "no Stage 1-2 history and lease share rose",
           burst         ~ "no Stage 1-2 history; lease-concentrated Stage 3",
           TRUE          ~ "lease share rose from Stage 1-2 to Stage 3"))

cand_nonfishing <- open_profiles |>
  filter(gfw_vessel_type %in% "FISHING") |>
  mutate(research = str_detect(names, "\\bNOAA\\b|^R/V|RESEARCH|SURVEY"),
         visit    = !flag %in% "USA" & days <= screen$fp_max_days &
                    h_day >= screen$fp_min_h_day,
         inconcl  = !flag %in% "USA" & gfw_gear_type %in% "INCONCLUSIVE") |>
  filter(research | visit | inconcl) |>
  mutate(list = "3_nonfishing_typed_fishing",
         reason = case_when(research ~ "research or survey name",
                            visit    ~ "short, intense foreign-flag visit",
                            TRUE     ~ "foreign flag with INCONCLUSIVE gear"))

cand_marker <- open_profiles |>
  filter(!mmsi_valid | battery_call_sign) |>
  mutate(list = "4_gear_marker",
         reason = if_else(battery_call_sign, "call sign holds a battery level",
                          "MMSI is not 9 digits"))

candidates <- bind_rows(cand_missed, cand_charter, cand_nonfishing, cand_marker) |>
  select(list, reason, MMSI, names, flag, call_sign, gfw_vessel_type,
         gfw_gear_type, vp_hours, s1, s2, s3, afe_hours, days, h_day,
         lease_share, lease_share_s12, lease_share_s3, s3_lease_first,
         s3_lease_last, first, last, any_of(c("burst", "shift")),
         starts_with("garfo_"), hull_is_uscg) |>
  arrange(list, desc(vp_hours))

candidates |>
  group_by(list) |>
  summarise(vessels = n(), vp_hours = sum(vp_hours), afe_hours = sum(afe_hours))
```

### How well does the charter screen find known charters?

The vessels Ørsted confirmed are the best available ground truth. The share
the screen flags is its recall; vessels it misses show what it cannot see.

```{r}
orsted_charters <- cw |>
  filter(Identified_By == "orsted_list", activity_excluded) |>
  distinct(MMSI)

screen_recall <- charter_screen(profiles) |>
  inner_join(orsted_charters, by = "MMSI")

screen_recall |>
  summarise(vessels = n(), flagged = sum(charter_flag),
            recall = round(mean(charter_flag), 2))

screen_recall |>
  filter(!charter_flag) |>
  select(MMSI, names, s1, s2, s3, lease_share, lease_share_s12, lease_share_s3)
```

### Gear review candidates

Fishing vessels are those GFW types FISHING plus the crosswalk's fishing hulls
(removal-list charter vessels and accepted missed fishing vessels). One of
them goes on the gear review sheet, until it has a row in
`vessel_gear.csv`, if:

- it is on the **removal list** -- every removal-list fishing vessel is
  reviewed, because GFW's gear can be wrong even when it is resolved (e.g.
  CAILYN AND MAREN, a lobster trap boat GFW labels `POLE_AND_LINE`);
- GFW's gear is **unresolved** (`FISHING`, `INCONCLUSIVE`, `OTHER`, ...);
- or a unique GARFO match's permits **disagree** with GFW's gear class.

```{r}
removal_mmsi <- read_chr(paths$removals)$MMSI

fishing_mmsi <- union(profiles$MMSI[profiles$gfw_vessel_type %in% "FISHING"],
                      cw$MMSI[cw$hull_class == "fishing"]) |>
  setdiff(cw$MMSI[cw$hull_class != "fishing"])

gear_candidates <- profiles |>
  filter(MMSI %in% fishing_mmsi, !MMSI %in% gear_inputs$MMSI) |>
  mutate(gfw_class       = gear_class_of(gfw_gear_type),
         garfo_hint      = if_else(garfo_match == "unique_years", garfo_gear_hint, NA_character_),
         on_removal_list = MMSI %in% removal_mmsi,
         unresolved      = gfw_class == "UNRESOLVED",
         garfo_conflict  = !unresolved & !is.na(garfo_hint) & garfo_hint != gfw_class) |>
  filter(on_removal_list | unresolved | garfo_conflict) |>
  mutate(gear_reason = str_remove(paste0(
    if_else(on_removal_list, "removal list; ", ""),
    if_else(unresolved, "GFW gear unresolved; ", ""),
    if_else(garfo_conflict, "GARFO permits disagree with GFW gear; ", "")), "; $"))

gear_candidates |>
  count(on_removal_list, unresolved, garfo_conflict, wt = afe_hours, name = "afe_hours") |>
  left_join(count(gear_candidates, on_removal_list, unresolved, garfo_conflict, name = "vessels"),
            by = c("on_removal_list", "unresolved", "garfo_conflict"))
```

## Phase 3: Gather evidence

### GFW vessel-identity API

Looks up every Vessel ID belonging to a candidate MMSI, on either review
sheet (the gear candidates need registry gear). Parsed results are
appended to CSVs in `data/raw/vessel_evidence/`, and every queried ID is
logged with the date and the GFW dataset version, so each ID is requested
once across runs. The registry records (`registryInfo`) are the independent
evidence: GFW's own label is derived from them, so a vessel typed OTHER whose
registry lists a fishing gear is a strong false-negative candidate, and a
vessel with no registry record at all has no evidence either way.

```{r}
evid_dir <- paths$evidence
dir.create(evid_dir, recursive = TRUE, showWarnings = FALSE)

ev_paths <- list(
  queried  = file.path(evid_dir, "gfw_identity_queried.csv"),
  registry = file.path(evid_dir, "gfw_identity_registry.csv"),
  combined = file.path(evid_dir, "gfw_identity_combined.csv"))

append_csv <- function(df, path) {
  if (nrow(df) > 0) write_csv(df, path, append = file.exists(path), na = "")
}

read_ev <- function(path, cols) {
  if (file.exists(path)) {
    read_chr(path)
  } else {
    as_tibble(set_names(rep(list(character()), length(cols)), cols))
  }
}

# character column from an API data frame; list columns collapsed with "|"
col_chr <- function(df, col) {
  if (!col %in% names(df)) return(rep(NA_character_, nrow(df)))
  x <- df[[col]]
  if (is.list(x)) map_chr(x, \(v) paste(unique(unlist(v)), collapse = "|"))
  else as.character(x)
}

queried_cols  <- c("vessel_id", "queried_on", "status", "dataset")
registry_cols <- c("MMSI", "source", "reg_mmsi", "reg_shipname", "reg_callsign",
                   "reg_gear", "reg_length_m", "reg_from", "reg_to", "reg_extra")
combined_cols <- c("MMSI", "vessel_id", "gear", "ship", "year_from", "year_to")

# Each API entry has an index; selfReportedInfo maps it to the MMSI(s) queried.
parse_identity <- function(r) {
  sr  <- r$selfReportedInfo
  idx <- tibble(index = as.numeric(sr$index), MMSI = col_chr(sr, "ssvid")) |>
    distinct()

  reg <- r$registryInfo
  registry <- if (is.data.frame(reg) && nrow(reg) > 0) {
    tibble(index        = as.numeric(reg$index),
           source       = col_chr(reg, "sourceCode"),
           reg_mmsi     = col_chr(reg, "ssvid"),
           reg_shipname = col_chr(reg, "shipname"),
           reg_callsign = col_chr(reg, "callsign"),
           reg_gear     = col_chr(reg, "geartypes"),
           reg_length_m = col_chr(reg, "lengthM"),
           reg_from     = col_chr(reg, "transmissionDateFrom"),
           reg_to       = col_chr(reg, "transmissionDateTo"),
           reg_extra    = col_chr(reg, "extraFields")) |>
      inner_join(idx, by = "index", relationship = "many-to-many") |>
      select(all_of(registry_cols)) |>
      distinct()
  } else {
    read_ev("", registry_cols)
  }

  cmb <- r$combinedSourcesInfo
  combined <- if (is.data.frame(cmb) && nrow(cmb) > 0) {
    tibble(index     = as.numeric(cmb$index),
           vessel_id = col_chr(cmb, "vesselId"),
           gear      = col_chr(cmb, "geartypes_name"),
           ship      = col_chr(cmb, "shiptypes_name"),
           year_from = col_chr(cmb, "shiptypes_yearFrom"),
           year_to   = col_chr(cmb, "shiptypes_yearTo")) |>
      inner_join(idx, by = "index", relationship = "many-to-many") |>
      select(all_of(combined_cols)) |>
      distinct()
  } else {
    read_ev("", combined_cols)
  }

  list(registry = registry, combined = combined,
       dataset = paste(unique(col_chr(r$dataset, "dataset")), collapse = "|"))
}
```

```{r}
id_lookup <- bind_rows(vp, afe) |>
  filter(!is.na(MMSI), !is.na(`Vessel ID`)) |>
  distinct(vessel_id = `Vessel ID`, MMSI)

queried <- read_ev(ev_paths$queried, queried_cols)

to_query <- id_lookup |>
  filter(MMSI %in% c(candidates$MMSI, gear_candidates$MMSI),
         !vessel_id %in% queried$vessel_id[queried$status %in% c("ok", "not_found")]) |>
  pull(vessel_id) |>
  unique()

message(length(to_query), " candidate Vessel IDs not yet queried")

if (run_gfw_api && length(to_query) > 0) {
  for (ids in split(to_query, ceiling(seq_along(to_query) / api_batch_size))) {
    res <- tryCatch(
      gfw_vessel_info(search_type = "id", ids = ids,
                      key = Sys.getenv("GFW_TOKEN"), quiet = TRUE),
      error = function(e) e)

    status  <- if (inherits(res, "error")) "error" else if (is.null(res)) "not_found" else "ok"
    dataset <- NA_character_

    if (status == "ok") {
      parsed <- parse_identity(res)
      append_csv(parsed$registry, ev_paths$registry)
      append_csv(parsed$combined, ev_paths$combined)
      dataset <- parsed$dataset
    }

    append_csv(tibble(vessel_id = ids, queried_on = as.character(Sys.Date()),
                      status = status, dataset = dataset),
               ev_paths$queried)
    Sys.sleep(0.5)
  }
}
```

```{r}
# summarise identity evidence per MMSI
queried  <- read_ev(ev_paths$queried, queried_cols)
registry <- read_ev(ev_paths$registry, registry_cols)
combined <- read_ev(ev_paths$combined, combined_cols)

count(queried, status, dataset)

queried_mmsi <- id_lookup |>
  filter(vessel_id %in% queried$vessel_id[queried$status %in% c("ok", "not_found")]) |>
  distinct(MMSI)

has_fishing_gear <- function(g) {
  map_lgl(str_split(coalesce(g, ""), fixed("|")), \(x) any(x %in% fishing_gear_codes))
}

gfw_evidence <- queried_mmsi |>
  left_join(registry |>
              group_by(MMSI) |>
              summarise(reg_records  = n(),
                        reg_sources  = paste(sort(unique(source)), collapse = "|"),
                        reg_gear     = paste(sort(unique(reg_gear)), collapse = "|"),
                        reg_length_m = suppressWarnings(
                          median(as.numeric(reg_length_m), na.rm = TRUE)),
                        .groups = "drop"),
            by = "MMSI") |>
  left_join(combined |>
              group_by(MMSI) |>
              summarise(gfw_combined = paste(unique(paste0(ship, " ", year_from, "-", year_to)),
                                             collapse = "; "),
                        .groups = "drop"),
            by = "MMSI") |>
  mutate(reg_records    = coalesce(reg_records, 0L),
         reg_length_m   = if_else(is.nan(reg_length_m), NA_real_, reg_length_m),
         reg_fishing    = reg_records > 0 & has_fishing_gear(reg_gear),
         reg_nonfishing = reg_records > 0 & !reg_fishing)

count(gfw_evidence, reg_records > 0, reg_fishing)
```

## Phase 4: The review sheet

### Suggested decisions

List 1 reaches the review sheet only when there is evidence: a GFW registry
record listing fishing gear, or a unique GARFO match with commercial or
for-hire permits. The rest of the pool cannot realistically be researched one
vessel at a time; its hours are reported below as an upper bound on what the
classification may still miss.

| Evidence | Suggestion |
|---|---|
| GFW registry lists fishing gear (list 1) | `fishing`, `confirmed` |
| Unique GARFO match, permit years overlap, length agrees (list 1) | `fishing`, `probable` -- confirm via the hull ID |
| Unique GARFO match, permit years overlap, no length to compare (list 1) | `fishing`, `possible` -- confirm via the hull ID |
| Charter burst with no prior history (list 2) | exclusion, `probable` |
| Lease-share rise only (list 2) | exclusion, `possible` |
| GFW registry lists non-fishing gear, no GARFO support (list 3) | `non_fishing`, `confirmed` |
| Name or behaviour only, or a GARFO match contradicts (list 3) | `non_fishing`, `possible` |
| Battery-level call sign (list 4) | `gear_marker`, `probable` |
| MMSI not 9 digits only (list 4) | `unknown`, `possible` -- check whether it is a vessel |

Charter windows are suggested from the first and last Stage 3 day with lease
activity; replace them with Ørsted's dates where available.

This sheet covers hull and activity only. Gear is set on the gear review
sheet (below); a missed fishing vessel you accept here appears there on the
next run, since GFW's gear for it is unresolved.

```{r}
evidence_cols <- c("list", "reason", "names", "flag", "call_sign", "vp_hours",
                   "s1", "s2", "s3", "afe_hours", "h_day", "lease_share",
                   "lease_share_s12", "lease_share_s3", "reg_records",
                   "reg_sources", "reg_gear", "gfw_combined", "reg_length_m",
                   "garfo_class", "garfo_hull_id", "hull_is_uscg",
                   "garfo_permits", "garfo_years", "garfo_length_ft",
                   "garfo_home_port", "garfo_permit_type", "garfo_gear_hint",
                   "garfo_oa_permits", "garfo_oa_directed", "garfo_categories",
                   "garfo_max_hulls_with_name", "garfo_max_mmsi_with_name")

assessed <- candidates |>
  left_join(gfw_evidence, by = "MMSI") |>
  mutate(
    across(c(reg_fishing, reg_nonfishing), \(x) coalesce(x, FALSE)),
    garfo_match = coalesce(garfo_match, "none"),

    # GARFO length (ft) against the GFW registry length (m), when both exist
    length_known = !is.na(reg_length_m) & !is.na(garfo_length_ft),
    length_agree = length_known &
      abs(garfo_length_ft * 0.3048 - reg_length_m) <= screen$length_tolerance * reg_length_m,
    garfo_class = case_when(
      garfo_match == "unique_years" & length_agree ~ "unique_years_length",
      garfo_match == "unique_years" & length_known ~ "length_conflict",
      TRUE                                         ~ garfo_match),
    # a recreational-only permit is not evidence of a fishing vessel
    garfo_support = garfo_class %in% c("unique_years_length", "unique_years") &
                    !garfo_permit_type %in% c("recreational only", "unmapped codes only"),
    has_evidence = reg_fishing | garfo_support)

review <- assessed |>
  # list 1 is reviewed only where there is evidence
  filter(list != "1_missed_fishing" | has_evidence) |>
  mutate(
    decision    = NA_character_,
    decided_on  = NA_character_,
    Vessel_Name = names,
    hull_class = case_when(
      list == "4_gear_marker" & str_detect(reason, "battery") ~ "gear_marker",
      list == "4_gear_marker"                                 ~ "unknown",
      list == "1_missed_fishing"                              ~ "fishing",
      list == "2_charter_activity"                            ~ "fishing",
      list == "3_nonfishing_typed_fishing"                    ~ "non_fishing"),
    activity_excluded = as.character(list == "2_charter_activity"),
    valid_from = if_else(list == "2_charter_activity",
                         as.character(s3_lease_first), NA_character_),
    valid_to   = if_else(list == "2_charter_activity",
                         as.character(s3_lease_last), NA_character_),
    Operation  = NA_character_,
    Confidence = case_when(
      list == "1_missed_fishing" & reg_fishing                          ~ "confirmed",
      list == "1_missed_fishing" & garfo_class == "unique_years_length" ~ "probable",
      list == "2_charter_activity" & burst                              ~ "probable",
      list == "3_nonfishing_typed_fishing" & reg_nonfishing & !garfo_support ~ "confirmed",
      list == "4_gear_marker" & str_detect(reason, "battery")           ~ "probable",
      TRUE                                                              ~ "possible"),
    Identified_By = case_when(
      list == "1_missed_fishing" & reg_fishing              ~ "gfw_registry",
      list == "1_missed_fishing" & garfo_support            ~ "garfo_name_match",
      list == "3_nonfishing_typed_fishing" & reg_nonfishing ~ "gfw_registry",
      list == "2_charter_activity"                          ~ "behavioural_screen",
      list == "4_gear_marker"                               ~ "mmsi_format_screen",
      TRUE                                                  ~ "screen"),
    evidence_ref = if_else(garfo_support,
                           paste0("GARFO hull ", garfo_hull_id, "; permit ", garfo_permits),
                           NA_character_),
    pull_date = as.character(pull_date),
    Notes = case_when(
      garfo_support & hull_is_uscg %in% TRUE ~ paste0(
        reason, ". To confirm: look up USCG official no. ", garfo_hull_id,
        " in CGMIX PSIX; its call sign should be ", coalesce(call_sign, "(none in AIS)"),
        ". GARFO permits: ", garfo_permit_type),
      garfo_support ~ paste0(
        reason, ". State-registered hull ", garfo_hull_id,
        "; cannot be linked to a call sign. GARFO permits: ", garfo_permit_type),
      list == "3_nonfishing_typed_fishing" & garfo_match == "unique_years" ~ paste0(
        reason, ". Note: unique GARFO permit match suggests a fishing vessel"),
      TRUE ~ reason)) |>
  # within each list: GFW registry evidence first, then GARFO with agreeing
  # length, then by hours
  arrange(list, desc(reg_fishing), desc(garfo_class == "unique_years_length"),
          desc(vp_hours)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(all_of(decision_cols), any_of(evidence_cols))
```

### Write the review sheet

The sheet is regenerated every run; anything you mark in it is collected into
`vessel_decisions.csv` at the start of the next run (Phase 1).

```{r}
dir.create(paths$review, recursive = TRUE, showWarnings = FALSE)
write_csv(review, review_path, na = "")

count(review, list)
```

### What is left unreviewed

List 1 vessels without evidence are not reviewed. Their hours bound what the
classification could still be missing; report them alongside the results.

```{r}
assessed |>
  filter(list == "1_missed_fishing") |>
  group_by(has_evidence) |>
  summarise(vessels = n(),
            vp_share  = round(sum(vp_hours) / total_vp, 4),
            afe_share = round(sum(afe_hours) / total_afe, 4),
            .groups = "drop")
```

### The gear review sheet

One row per gear candidate (Phase 2): removal-list fishing vessels first, then
the rest by fishing-effort hours. The suggested `gear_class` comes from a
unique GARFO match's permits if there is one, otherwise from the GFW registry
gear (after the API run), otherwise it is blank.

| GARFO permits | Suggested gear class |
|---|---|
| For-hire only | `POLE_AND_LINE` |
| For-hire plus `screen$oa_bundle_min` or more open-access commercial permits -- the bundle party boats typically hold | `POLE_AND_LINE`, but treat it as `possible` |
| For-hire plus fewer open-access commercial permits | blank -- the vessel may fish commercially on some trips (e.g. Handgear B, tilefish); the note lists the directed open-access permits |
| For-hire plus limited-access commercial permits | blank |
| Commercial: trap permits only (lobster trap areas, red crab, a trap limit) | `FIXED` |
| Commercial: dredge or trawl permits only (scallop, clam, squid trawl) | `MOBILE` |
| Anything else | blank |

`garfo_gear_hint` and `garfo_categories` are shown even when the GARFO match
is ambiguous, as information. A vessel gets one gear class per window, so a
dual-permitted vessel that alternates charter and commercial trips is
misclassified for some of its trips whichever class is chosen; Phase 6
reports these vessels as a separate group.

```{r}
# a single analysis gear class from the GFW registry gear codes, if they agree
reg_gear_class <- function(g) {
  map_chr(str_split(coalesce(g, ""), fixed("|")), \(x) {
    cls <- setdiff(unique(gear_class_of(x[x != ""])), "UNRESOLVED")
    if (length(cls) == 1) cls else NA_character_
  })
}

gear_evidence_cols <- c("gear_reason", "gfw_class", "garfo_match", "garfo_gear_hint",
                        "garfo_permit_type", "garfo_mmsi_sequence", "garfo_categories",
                        "garfo_hull_id", "reg_records", "reg_gear", "names",
                        "vp_hours", "afe_hours", "s1", "s2", "s3")

gear_review <- gear_candidates |>
  left_join(select(gfw_evidence, MMSI, reg_records, reg_gear), by = "MMSI") |>
  mutate(
    reg_hint    = reg_gear_class(reg_gear),
    decision    = NA_character_,
    decided_on  = NA_character_,
    valid_from  = NA_character_,
    valid_to    = NA_character_,
    Vessel_Name = names,
    gear_class  = coalesce(garfo_hint, reg_hint),
    basis = case_when(!is.na(garfo_hint) ~ "GARFO permit",
                      !is.na(reg_hint)   ~ "GFW registry",
                      TRUE               ~ NA_character_),
    evidence_ref = if_else(!is.na(garfo_hint),
                           paste0("GARFO hull ", garfo_hull_id, "; permit ", garfo_permits),
                           NA_character_),
    pull_date = as.character(pull_date),
    Notes = paste0(
      gear_reason,
      if_else(!is.na(garfo_hint) & !is.na(reg_hint) & garfo_hint != reg_hint,
              paste0(". GFW registry gear suggests ", reg_hint, " instead"), ""),
      case_when(
        garfo_oa_bundle %in% TRUE ~ paste0(
          ". Holds ", garfo_n_oa, " open-access commercial permits, a typical party-boat",
          " bundle; POLE_AND_LINE suggested -- treat it as possible"),
        garfo_permit_type %in% "for-hire + open-access commercial" ~ paste0(
          ". Also holds open-access commercial permits (directed: ",
          if_else(garfo_oa_directed %in% "", "none", garfo_oa_directed),
          "); may fish commercially on some trips"),
        TRUE ~ ""))) |>
  arrange(desc(on_removal_list), desc(afe_hours)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(all_of(gear_cols), any_of(gear_evidence_cols))

write_csv(gear_review, gear_review_path, na = "")

count(gear_review, basis, gear_class)
```

## Phase 5: Classify every record

Each record is matched to at most one crosswalk row by MMSI and date. Records
with no row keep GFW's label. Three analysis sets follow:

| Set | Fishing hulls GFW missed that are added | Non-fishing hulls and charter activity removed |
|---|---|---|
| **Strict** | none | `confirmed` only |
| **Primary** | `confirmed` | `confirmed` and `probable` |
| **Inclusive** | `confirmed`, `probable`, `possible` | `confirmed` and `probable` |

Gear markers are removed from every set. The sets differ along two axes, so
they are not strictly nested: Strict adds nothing but also removes the least.
`possible` exclusions (e.g. ESS Pride) are removed from no set; list them as
a stated sensitivity rather than as a filter.

Each record's `gear_class` is the accepted correction in `vessel_gear.csv`
whose window contains its date, otherwise GFW's `Gear Type` through the gear
map; `gear_source` says which.

```{r}
classify <- function(df, cw, gear) {
  cw_join <- cw |>
    transmute(MMSI,
              cw_from = coalesce(valid_from, lo_date),
              cw_to   = coalesce(valid_to, hi_date),
              cw_hull = hull_class,
              cw_excluded = activity_excluded,
              cw_confidence = Confidence,
              cw_identified_by = Identified_By)

  gear_join <- gear |>
    transmute(MMSI,
              g_from = coalesce(valid_from, lo_date),
              g_to   = coalesce(valid_to, hi_date),
              gear_override = gear_class)

  out <- df |>
    left_join(cw_join, by = join_by(MMSI, between(date, cw_from, cw_to))) |>
    left_join(gear_join, by = join_by(MMSI, between(date, g_from, g_to)))

  stopifnot("a record matched more than one crosswalk or gear row" = nrow(out) == nrow(df))

  out |>
    mutate(
      gfw_fishing = `Vessel Type` %in% "FISHING",
      conf_cp     = cw_confidence %in% c("confirmed", "probable"),
      is_marker   = cw_hull %in% "gear_marker",
      nonfish_c   = cw_hull %in% "non_fishing" & cw_confidence %in% "confirmed",
      nonfish_cp  = cw_hull %in% "non_fishing" & conf_cp,
      added_c     = cw_hull %in% "fishing" & !gfw_fishing & cw_confidence %in% "confirmed",
      added_any   = cw_hull %in% "fishing" & !gfw_fishing,
      excl_c      = cw_excluded %in% TRUE & cw_confidence %in% "confirmed",
      excl_cp     = cw_excluded %in% TRUE & conf_cp,
      in_strict    = gfw_fishing & !is_marker & !nonfish_c & !excl_c,
      in_primary   = (gfw_fishing | added_c) & !is_marker & !nonfish_cp & !excl_cp,
      in_inclusive = (gfw_fishing | added_any) & !is_marker & !nonfish_cp & !excl_cp,
      gear_class  = coalesce(gear_override, gear_class_of(`Gear Type`)),
      gear_source = if_else(is.na(gear_override), "gfw", "vessel_gear")) |>
    select(-cw_from, -cw_to, -g_from, -g_to, -gear_override, -conf_cp)
}

vp_cls  <- classify(vp, cw, gear)
afe_cls <- classify(afe, cw, gear)
```

```{r}
set_cols <- c("in_strict", "in_primary", "in_inclusive")

bind_rows(VP = vp_cls, AFE = afe_cls, .id = "dataset") |>
  group_by(dataset) |>
  summarise(across(all_of(set_cols), \(x) sum(hours[x])),
            total = sum(hours), .groups = "drop")
```

```{r}
write_csv(vp_cls,  file.path(paths$processed_gfw, "gfw_vp_id_classified.csv"),  na = "")
write_csv(afe_cls, file.path(paths$processed_gfw, "gfw_afe_id_classified.csv"), na = "")
```

## Phase 6: Account for what is left

### Stage totals and Stage 2 → 3 change by set

```{r}
to_long <- function(df) {
  df |>
    pivot_longer(all_of(set_cols), names_to = "set", values_to = "keep") |>
    filter(keep) |>
    mutate(set = factor(str_remove(set, "in_"),
                        levels = c("strict", "primary", "inclusive")))
}

stage_by_set <- bind_rows(VP = to_long(vp_cls), AFE = to_long(afe_cls), .id = "dataset") |>
  group_by(dataset, set, `Development Stage`) |>
  summarise(hours = sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop")

stage_by_set

stage_by_set |>
  select(-n_mmsi) |>
  pivot_wider(names_from = `Development Stage`, values_from = hours) |>
  mutate(change_s2_s3_pct = round(100 * (`Stage 3` - `Stage 2`) / `Stage 2`, 1))
```

### Fishing effort missing for added vessels

GFW computes AFE only for vessels it treats as fishing, so hulls added in
Primary or Inclusive have presence hours but almost no effort hours. Report
this as a known undercount. The imputation below applies each stage's
effort-to-presence ratio (from Strict-set vessels, by gear class) to the added
vessels' presence hours; use it as a sensitivity check only.

```{r}
added <- vp_cls |> filter(added_any, in_inclusive)

ratio <- inner_join(
  vp_cls  |> filter(in_strict) |> group_by(`Development Stage`, gear_class) |>
    summarise(vp = sum(hours), .groups = "drop"),
  afe_cls |> filter(in_strict) |> group_by(`Development Stage`, gear_class) |>
    summarise(afe = sum(hours), .groups = "drop"),
  by = c("Development Stage", "gear_class")) |>
  mutate(afe_per_vp = afe / vp)

added |>
  group_by(`Development Stage`, gear_class,
           set = if_else(added_c, "primary + inclusive", "inclusive only")) |>
  summarise(vp_hours = sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop") |>
  left_join(afe_cls |> filter(added_any) |>
              group_by(`Development Stage`, gear_class) |>
              summarise(afe_hours_reported = sum(hours), .groups = "drop"),
            by = c("Development Stage", "gear_class")) |>
  left_join(select(ratio, `Development Stage`, gear_class, afe_per_vp),
            by = c("Development Stage", "gear_class")) |>
  mutate(afe_hours_reported = coalesce(afe_hours_reported, 0),
         afe_hours_imputed  = vp_hours * afe_per_vp)
```

### Dual-permitted vessels

Vessels whose GARFO permits combine for-hire with commercial permits
(`commercial + for-hire`, `for-hire + open-access commercial`) may fish both
ways, on different trips. A single gear class cannot represent them, so report
their hours as a separate group and check whether the Stage 2 → 3 change holds
with them excluded. Only unique, year-overlapping GARFO matches are grouped;
everything else is `no unique GARFO match`.

```{r}
permit_group <- profiles |>
  transmute(MMSI,
            permit_group = if_else(garfo_match == "unique_years",
                                   garfo_permit_type, NA_character_))

bind_rows(VP = to_long(vp_cls), AFE = to_long(afe_cls), .id = "dataset") |>
  left_join(permit_group, by = "MMSI") |>
  mutate(permit_group = coalesce(permit_group, "no unique GARFO match")) |>
  group_by(dataset, set, permit_group, `Development Stage`) |>
  summarise(hours = sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop") |>
  pivot_wider(names_from = `Development Stage`, values_from = c(hours, n_mmsi),
              values_fill = 0)
```

### Cell × stage tables by set

Same zero-filled layout as `gfw_fishnet_grid.qmd`, with a `set` column, ready
for the spatial analysis.

```{r}
cell_stage_sets <- function(df, value) {
  to_long(df) |>
    group_by(cell_id, `Development Stage`, set) |>
    summarise("{value}" := sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop") |>
    complete(cell_id = grid$cell_id, `Development Stage`, set,
             fill = set_names(list(0, 0L), c(value, "n_mmsi")))
}

vp_cell_stage_sets  <- cell_stage_sets(vp_cls,  "vp_hours")
afe_cell_stage_sets <- cell_stage_sets(afe_cls, "afe_hours")

stopifnot(nrow(vp_cell_stage_sets) == nrow(grid) * nrow(dev_stages) * 3)

write_csv(vp_cell_stage_sets,  file.path(paths$interim_gfw, "gfw_vp_cell_stage_sets.csv"))
write_csv(afe_cell_stage_sets, file.path(paths$interim_gfw, "gfw_afe_cell_stage_sets.csv"))
```

## Phase 7: Repeatability checks

### GFW label drift

GFW's labels change between pulls (CYGNUS was typed FISHING/INCONCLUSIVE in an
earlier pull and CARGO in the 2026-10-06 pull). Any decided vessel whose
current label differs from the one recorded when the decision was made is
listed here for a second look.

```{r}
decisions |>
  filter(!is.na(gfw_vessel_type)) |>
  distinct(MMSI, Vessel_Name, gfw_vessel_type, gfw_gear_type, pull_date) |>
  left_join(gfw_labels, by = "MMSI", suffix = c("_recorded", "_current")) |>
  filter(is.na(gfw_vessel_type_current) |
           gfw_vessel_type_recorded != gfw_vessel_type_current |
           gfw_gear_type_recorded != gfw_gear_type_current)
```

### Crosswalk summary for the methods

```{r}
touched <- function(df) {
  df |>
    filter(!is.na(cw_hull)) |>
    group_by(cw_hull, cw_excluded, cw_confidence) |>
    summarise(hours = sum(hours), .groups = "drop")
}

count(cw, hull_class, activity_excluded, Confidence, name = "rows")

full_join(touched(vp_cls), touched(afe_cls),
          by = c("cw_hull", "cw_excluded", "cw_confidence"),
          suffix = c("_vp", "_afe")) |>
  mutate(share_vp  = round(hours_vp  / total_vp,  4),
         share_afe = round(hours_afe / total_afe, 4))
```

### Gear summary for the methods

How much of each set's fishing effort has a resolved gear class, and how much
of it comes from your gear corrections rather than GFW.

```{r}
to_long(afe_cls) |>
  group_by(set, gear_class, gear_source) |>
  summarise(afe_hours = round(sum(hours)), .groups = "drop") |>
  group_by(set) |>
  mutate(share = round(afe_hours / sum(afe_hours), 3)) |>
  ungroup()
```
