---
title: "MM GFW Vessel Classification: False Positives and False Negatives"
author: "Malachy McCaffrey"
date: today
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Identify and account for vessels that GFW misclassifies in the raw VP and AFE
pulls, following the hybrid approach in `r/scratchpad/vessel_filtering_plan.qmd`:

- an **MMSI crosswalk** (`references/vessel_crosswalk.csv`) is the single record
  of every classification decision
- a **screen** of the raw data produces ranked review lists
- the **GFW vessel-identity API** and **external registries** supply evidence
- **three analysis sets** (Strict / Primary / Inclusive) carry what stays
  uncertain into the results

Two separate problems are handled:

- **Hull identity** -- is the vessel a fishing vessel at all? GFW types some
  fishing vessels OTHER, GEAR or PASSENGER (false negatives) and some
  non-fishing vessels FISHING (false positives).
- **Activity** -- is a fishing vessel fishing at this moment? Charter guard and
  survey work by fishing hulls is classed as fishing by GFW (false positives).
  This depends on dates, so crosswalk rows carry a validity window.

## How to use this workflow

The workflow is iterative. Each pass:

1. Render (or run) this qmd. The first run seeds the crosswalk from
   `references/vessel_removals.csv`.
2. Open `data/processed/vessel_review/vessel_review.csv`. For each candidate,
   set `decision` to `accept` or `reject`, editing the `final_*` columns first
   if the suggestion is wrong (e.g. the charter window dates).
3. Re-run. Accepted rows are appended to the crosswalk, and the review file
   keeps every decision so rejected vessels are not asked about again.
4. Repeat until the unreviewed share of hours in every list is below the
   review cut-off (Phase 4).

`run_gfw_api` is `FALSE` by default so a render never makes API calls by
accident. Set it to `TRUE` for passes that should look up new candidates;
results are cached, so each Vessel ID is queried once.

## Libraries

```{r}
library(tidyverse)
library(sf)
library(gfwr)
```

## Parameters

```{r}
pull_date   <- as.Date("2026-10-06") # date of the raw pull (gfw_vp_afe_datapull.qmd)
run_gfw_api <- FALSE                 # TRUE to query uncached candidates

paths <- list(
  vp        = "data/raw/gfw/gfw_vp_id.csv",
  afe       = "data/raw/gfw/gfw_afe_id.csv",
  removals  = "references/vessel_removals.csv",
  crosswalk = "references/vessel_crosswalk.csv",
  grid      = "data/sf/fishnet/gfw_grid_utm19n.gpkg",
  leases    = c("data/sf/owf/SFW.shp", "data/sf/owf/RWF.shp", "data/sf/owf/SRW.shp"),
  evidence  = "data/raw/vessel_evidence",     # API cache and registry extracts
  review    = "data/processed/vessel_review", # review lists and decisions
  out_gfw   = "data/processed/gfw",           # classified records
  out_grid  = "data/interim/gfw"             # cell x stage x set tables
)

# screening thresholds -- record any change in the methods
screen <- list(
  min_s3_hours    = 50,   # charter screen: minimum Stage 3 VP hours
  min_lease_share = 0.25, # charter screen: share of hours inside SFW/RWF/SRW
  max_prior_hours = 0,    # charter screen: Stage 1-2 hours for a "no history" burst
  min_lease_shift = 0.25, # charter screen: rise in lease share, Stage 1-2 to 3
  fp_max_days     = 30,   # FP screen: short foreign-flag visit, active days
  fp_min_h_day    = 12,   # FP screen: hours per active day
  review_cutoff   = 0.01  # stop reviewing when unreviewed hours < 1% of dataset
)

api_batch_size <- 20 # Vessel IDs per API request
```

## Read raw data

```{r}
# Stage boundaries mirror gfw_vp_afe_datapull.qmd and gfw_fishnet_grid.qmd --
# keep all three in sync.
stage_of <- function(ym) {
  case_when(
    between(ym, as.Date("2016-01-01"), as.Date("2019-08-01")) ~ "Stage 1",
    between(ym, as.Date("2019-09-01"), as.Date("2023-03-01")) ~ "Stage 2",
    between(ym, as.Date("2023-04-01"), as.Date("2026-09-01")) ~ "Stage 3")
}

read_gfw <- function(path, hours_col) {
  read_csv(path, show_col_types = FALSE,
           col_types = cols(.default = col_character())) |>
    mutate(hours = as.numeric(.data[[hours_col]]),
           date  = as.Date(`Time Range`),
           `Year Month` = floor_date(date, "month"),
           `Development Stage` = stage_of(`Year Month`),
           cell_id = sprintf("%d_%d",
                             as.integer(round(as.numeric(Lon) / 0.01)),
                             as.integer(round(as.numeric(Lat) / 0.01))))
}

vp  <- read_gfw(paths$vp,  "Vessel Presence Hours")
afe <- read_gfw(paths$afe, "Apparent Fishing Hours")

stopifnot("records with no development stage" =
            !anyNA(vp$`Development Stage`) && !anyNA(afe$`Development Stage`))
```

```{r}
# GFW labels per MMSI as of this pull (most frequent, in case of a split)
gfw_labels <- bind_rows(vp, afe) |>
  filter(!is.na(MMSI)) |>
  count(MMSI, `Vessel Type`, `Gear Type`) |>
  group_by(MMSI) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(MMSI, gfw_vessel_type = `Vessel Type`, gfw_gear_type = `Gear Type`)

# GFW gear type -> analysis gear class (references/data_dictionary.md)
gear_class_map <- c(
  TRAWLERS = "MOBILE", DREDGE_FISHING = "MOBILE", OTHER_PURSE_SEINES = "MOBILE",
  TUNA_PURSE_SEINES = "MOBILE", PURSE_SEINES = "MOBILE", TROLLERS = "MOBILE",
  DRIFTING_LONGLINES = "MOBILE",
  SET_GILLNETS = "FIXED", POTS_AND_TRAPS = "FIXED", SET_LONGLINES = "FIXED",
  FIXED_GEAR = "FIXED",
  POLE_AND_LINE = "POLE_AND_LINE")

gear_class_of <- function(g) coalesce(unname(gear_class_map[g]), "UNRESOLVED")

fishing_gear_codes <- c(names(gear_class_map), "FISHING", "INCONCLUSIVE")
```

```{r}
# lease flag per grid cell (cell centre inside SFW, RWF or SRW)
grid <- st_read(paths$grid, quiet = TRUE)

leases <- paths$leases |>
  map(\(p) st_read(p, quiet = TRUE) |> st_transform(st_crs(grid)) |> st_union()) |>
  reduce(st_union)

cell_lease <- tibble(
  cell_id  = grid$cell_id,
  in_lease = lengths(st_intersects(st_centroid(st_geometry(grid)), leases)) > 0)

stopifnot("records outside the fishnet grid" =
            all(c(vp$cell_id, afe$cell_id) %in% cell_lease$cell_id))
```

## Phase 1: The MMSI crosswalk

One row per MMSI per validity period. A record is matched to the row whose
`valid_from`–`valid_to` window contains its date (blank = open-ended). Records
with no matching row keep GFW's label. Windows for one MMSI must not overlap,
so a fishing hull GFW missed that also took a charter needs two rows: one
before the charter (`activity_excluded = FALSE`) and one for the charter
(`TRUE`).

| Column | Values |
|---|---|
| `MMSI` | 9 digits (any format for `gear_marker`) |
| `valid_from`, `valid_to` | ISO dates; blank = open-ended |
| `hull_class` | `fishing` / `non_fishing` / `gear_marker` / `unknown` |
| `gear_class` | `MOBILE` / `FIXED` / `POLE_AND_LINE` / `UNRESOLVED`; blank = use GFW gear |
| `activity_excluded` | `TRUE` = records in the window are not fishing (charter work) |
| `Confidence` | `confirmed` / `probable` / `possible` |
| `Identified_By`, `evidence_ref`, `Notes` | Where the decision came from |
| `gfw_vessel_type`, `gfw_gear_type`, `pull_date` | GFW's labels when the row was written |

### Seed from the removal list

Runs only when the crosswalk does not exist yet, so manual edits are never
overwritten. Charter vessels (`Scope == "stages_2_3"`) become fishing hulls
with an activity exclusion from the start of Stage 2; replace that date with
the real charter window wherever Ørsted's records give one. `all_stages`
vessels become non-fishing hulls. `vessel_removals.csv` itself is not changed.

```{r}
cw_cols <- c("MMSI", "valid_from", "valid_to", "Vessel_Name", "hull_class",
             "gear_class", "activity_excluded", "Operation", "Identified_By",
             "Confidence", "evidence_ref", "gfw_vessel_type", "gfw_gear_type",
             "pull_date", "Notes")

if (!file.exists(paths$crosswalk)) {
  read_csv(paths$removals, col_types = cols(.default = col_character())) |>
    transmute(MMSI, Vessel_Name,
              valid_from = if_else(Scope == "stages_2_3", "2019-09-01", NA_character_),
              valid_to   = NA_character_,
              hull_class = if_else(Scope == "stages_2_3", "fishing", "non_fishing"),
              gear_class = NA_character_,
              activity_excluded = Scope == "stages_2_3",
              Operation, Identified_By, Confidence,
              evidence_ref = NA_character_,
              pull_date = as.character(pull_date),
              Notes) |>
    left_join(gfw_labels, by = "MMSI") |>
    select(all_of(cw_cols)) |>
    write_csv(paths$crosswalk, na = "")

  message("Seeded ", paths$crosswalk, " from ", paths$removals)
}
```

### Read and validate

```{r}
lo_date <- as.Date("1900-01-01") # stand-ins for open-ended windows
hi_date <- as.Date("2100-12-31")

read_crosswalk <- function(path) {
  cw <- read_csv(path, col_types = cols(.default = col_character()),
                 na = c("", "NA"))

  missing_cols <- setdiff(cw_cols, names(cw))
  if (length(missing_cols) > 0) {
    stop("crosswalk is missing columns: ", paste(missing_cols, collapse = ", "),
         call. = FALSE)
  }

  cw <- cw |>
    mutate(valid_from = as.Date(valid_from),
           valid_to   = as.Date(valid_to),
           activity_excluded = as.logical(activity_excluded))

  overlaps <- cw |>
    mutate(f = coalesce(valid_from, lo_date), t = coalesce(valid_to, hi_date)) |>
    arrange(MMSI, f) |>
    group_by(MMSI) |>
    filter(lead(f) <= t | lag(t) >= f) |>
    pull(MMSI) |>
    unique()

  problems <- c(
    if (any(!cw$hull_class %in% c("fishing", "non_fishing", "gear_marker", "unknown")))
      "unknown hull_class",
    if (any(!cw$Confidence %in% c("confirmed", "probable", "possible")))
      "unknown Confidence",
    if (any(!is.na(cw$gear_class) &
            !cw$gear_class %in% c("MOBILE", "FIXED", "POLE_AND_LINE", "UNRESOLVED")))
      "unknown gear_class",
    if (anyNA(cw$activity_excluded))
      "activity_excluded must be TRUE or FALSE",
    if (any(cw$hull_class != "gear_marker" & !grepl("^[0-9]{9}$", cw$MMSI)))
      "MMSI is not 9 digits (allowed only for gear_marker)",
    if (any(cw$valid_from > cw$valid_to, na.rm = TRUE))
      "valid_from after valid_to",
    if (length(overlaps) > 0)
      paste0("overlapping windows for MMSI: ", paste(overlaps, collapse = ", "))
  )

  if (length(problems) > 0) {
    stop("Invalid ", path, ":\n  - ", paste(problems, collapse = "\n  - "),
         call. = FALSE)
  }

  cw
}

cw <- read_crosswalk(paths$crosswalk)

count(cw, hull_class, activity_excluded, Confidence)
```

## Phase 2: Screen the raw data

### Per-MMSI profile

```{r}
profile_mmsi <- function(vp, afe, cell_lease) {
  v <- vp |> filter(!is.na(MMSI)) |> left_join(cell_lease, by = "cell_id")
  a <- afe |> filter(!is.na(MMSI))

  stage_hours <- function(h, stage, s) sum(h[stage == s])

  pv <- v |>
    group_by(MMSI) |>
    summarise(
      names     = paste(sort(unique(na.omit(`Vessel Name`))), collapse = " / "),
      flag      = first(na.omit(Flag)),
      call_sign = first(na.omit(CallSign)),
      vp_hours  = sum(hours),
      s1 = stage_hours(hours, `Development Stage`, "Stage 1"),
      s2 = stage_hours(hours, `Development Stage`, "Stage 2"),
      s3 = stage_hours(hours, `Development Stage`, "Stage 3"),
      days  = n_distinct(date),
      h_day = vp_hours / days,
      lease_share     = sum(hours[in_lease]) / vp_hours,
      lease_share_s12 = if (s1 + s2 > 0)
        sum(hours[in_lease & `Development Stage` != "Stage 3"]) / (s1 + s2) else NA_real_,
      lease_share_s3  = if (s3 > 0)
        sum(hours[in_lease & `Development Stage` == "Stage 3"]) / s3 else NA_real_,
      s3_lease_first = suppressWarnings(min(date[in_lease & `Development Stage` == "Stage 3"])),
      s3_lease_last  = suppressWarnings(max(date[in_lease & `Development Stage` == "Stage 3"])),
      first = min(date),
      last  = max(date),
      .groups = "drop") |>
    mutate(across(c(s3_lease_first, s3_lease_last), \(x) if_else(is.finite(x), x, NA)))

  pa <- a |>
    group_by(MMSI) |>
    summarise(afe_hours = sum(hours),
              afe_s3 = sum(hours[`Development Stage` == "Stage 3"]),
              .groups = "drop")

  pv |>
    left_join(pa, by = "MMSI") |>
    left_join(gfw_labels, by = "MMSI") |>
    mutate(afe_hours = coalesce(afe_hours, 0),
           afe_s3    = coalesce(afe_s3, 0),
           mmsi_valid = grepl("^[0-9]{9}$", MMSI),
           battery_call_sign = str_detect(coalesce(call_sign, ""), "%"))
}

profiles <- profile_mmsi(vp, afe, cell_lease)

nrow(profiles)
```

### Candidate lists

Each list excludes MMSIs already in the crosswalk.

1. **Missed fishing vessels** -- typed other than FISHING, US-flag, with a call
   sign (the only vessels US registries can confirm).
2. **Charter activity** -- typed FISHING with a lease-concentrated Stage 3:
   either no Stage 1–2 history (a burst), or a lease share that rose from
   Stages 1–2 to Stage 3 (catches established vessels that took a charter, the
   *Virginia Wave* blind spot).
3. **Non-fishing vessels typed FISHING** -- research or survey names, short
   intense foreign-flag visits, or foreign-flag `INCONCLUSIVE` gear.
4. **Gear markers and malformed MMSIs** -- call signs holding battery levels
   (e.g. ONWA33NT, likely AIS net buoys), or MMSIs that are not 9 digits
   (some are vessels with misconfigured transponders, e.g. ROYAL with call sign
   WDD3653, so these are suggested as `unknown` for review).

```{r}
charter_screen <- function(p) {
  p |>
    mutate(burst = s1 + s2 <= screen$max_prior_hours &
                   lease_share >= screen$min_lease_share,
           shift = !is.na(lease_share_s12) & !is.na(lease_share_s3) &
                   lease_share_s3 - lease_share_s12 >= screen$min_lease_shift,
           charter_flag = s3 >= screen$min_s3_hours & (burst | shift))
}

not_listed <- !profiles$MMSI %in% cw$MMSI

cand_missed <- profiles[not_listed, ] |>
  filter(!gfw_vessel_type %in% "FISHING", mmsi_valid,
         flag %in% "USA", !is.na(call_sign)) |>
  mutate(list = "1_missed_fishing",
         reason = "typed non-FISHING; US flag with call sign")

cand_charter <- charter_screen(profiles[not_listed, ]) |>
  filter(gfw_vessel_type %in% "FISHING", charter_flag) |>
  mutate(list = "2_charter_activity",
         reason = case_when(
           burst & shift ~ "no Stage 1-2 history and lease share rose",
           burst         ~ "no Stage 1-2 history; lease-concentrated Stage 3",
           TRUE          ~ "lease share rose from Stage 1-2 to Stage 3"))

cand_nonfishing <- profiles[not_listed, ] |>
  filter(gfw_vessel_type %in% "FISHING") |>
  mutate(research = str_detect(names, "\\bNOAA\\b|^R/V|RESEARCH|SURVEY"),
         visit    = !flag %in% "USA" & days <= screen$fp_max_days &
                    h_day >= screen$fp_min_h_day,
         inconcl  = !flag %in% "USA" & gfw_gear_type %in% "INCONCLUSIVE") |>
  filter(research | visit | inconcl) |>
  mutate(list = "3_nonfishing_typed_fishing",
         reason = case_when(research ~ "research or survey name",
                            visit    ~ "short, intense foreign-flag visit",
                            TRUE     ~ "foreign flag with INCONCLUSIVE gear"))

cand_marker <- profiles[not_listed, ] |>
  filter(!mmsi_valid | battery_call_sign) |>
  mutate(list = "4_gear_marker",
         reason = if_else(battery_call_sign, "call sign holds a battery level",
                          "MMSI is not 9 digits"))

candidates <- bind_rows(cand_missed, cand_charter, cand_nonfishing, cand_marker) |>
  select(list, reason, MMSI, names, flag, call_sign, gfw_vessel_type,
         gfw_gear_type, vp_hours, s1, s2, s3, afe_hours, days, h_day,
         lease_share, lease_share_s12, lease_share_s3, s3_lease_first,
         s3_lease_last, first, last, any_of(c("burst", "shift"))) |>
  arrange(list, desc(vp_hours))

candidates |>
  group_by(list) |>
  summarise(vessels = n(), vp_hours = sum(vp_hours), afe_hours = sum(afe_hours))
```

### How well does the charter screen find known charters?

The 42 vessels Ørsted confirmed are the best available ground truth. The share
the screen flags is its recall; vessels it misses show what it cannot see.

```{r}
orsted_charters <- cw |>
  filter(Identified_By == "orsted_list", activity_excluded) |>
  distinct(MMSI)

screen_recall <- charter_screen(profiles) |>
  inner_join(orsted_charters, by = "MMSI")

screen_recall |>
  summarise(vessels = n(), flagged = sum(charter_flag),
            recall = round(mean(charter_flag), 2))

screen_recall |>
  filter(!charter_flag) |>
  select(MMSI, names, s1, s2, s3, lease_share, lease_share_s12, lease_share_s3)
```

## Phase 3: Gather evidence

### GFW vessel-identity API

Looks up every Vessel ID belonging to a candidate MMSI. Parsed results are
appended to CSVs in `data/raw/vessel_evidence/`, and every queried ID is
logged, so each ID is requested once across runs. The registry records
(`registryInfo`) are the independent evidence: GFW's own label is derived from
them, so a vessel typed OTHER whose registry lists a fishing gear is a strong
false-negative candidate, and a vessel with no registry record at all has no
evidence either way.

```{r}
evid_dir <- paths$evidence
dir.create(evid_dir, recursive = TRUE, showWarnings = FALSE)

ev_paths <- list(
  queried  = file.path(evid_dir, "gfw_identity_queried.csv"),
  registry = file.path(evid_dir, "gfw_identity_registry.csv"),
  combined = file.path(evid_dir, "gfw_identity_combined.csv"))

append_csv <- function(df, path) {
  if (nrow(df) > 0) write_csv(df, path, append = file.exists(path), na = "")
}

read_ev <- function(path, cols) {
  if (file.exists(path)) {
    read_csv(path, col_types = cols(.default = col_character()), na = c("", "NA"))
  } else {
    as_tibble(set_names(rep(list(character()), length(cols)), cols))
  }
}

# character column from an API data frame; list columns collapsed with "|"
col_chr <- function(df, col) {
  if (!col %in% names(df)) return(rep(NA_character_, nrow(df)))
  x <- df[[col]]
  if (is.list(x)) map_chr(x, \(v) paste(unique(unlist(v)), collapse = "|"))
  else as.character(x)
}

registry_cols <- c("MMSI", "source", "reg_mmsi", "reg_shipname", "reg_callsign",
                   "reg_gear", "reg_from", "reg_to")
combined_cols <- c("MMSI", "vessel_id", "gear", "ship", "year_from", "year_to")

# Each API entry has an index; selfReportedInfo maps it to the MMSI(s) queried.
parse_identity <- function(r) {
  sr  <- r$selfReportedInfo
  idx <- tibble(index = as.numeric(sr$index), MMSI = col_chr(sr, "ssvid")) |>
    distinct()

  reg <- r$registryInfo
  registry <- if (is.data.frame(reg) && nrow(reg) > 0) {
    tibble(index        = as.numeric(reg$index),
           source       = col_chr(reg, "sourceCode"),
           reg_mmsi     = col_chr(reg, "ssvid"),
           reg_shipname = col_chr(reg, "shipname"),
           reg_callsign = col_chr(reg, "callsign"),
           reg_gear     = col_chr(reg, "geartypes"),
           reg_from     = col_chr(reg, "transmissionDateFrom"),
           reg_to       = col_chr(reg, "transmissionDateTo")) |>
      inner_join(idx, by = "index", relationship = "many-to-many") |>
      select(all_of(registry_cols)) |>
      distinct()
  } else {
    read_ev("", registry_cols)
  }

  cmb <- r$combinedSourcesInfo
  combined <- if (is.data.frame(cmb) && nrow(cmb) > 0) {
    tibble(index     = as.numeric(cmb$index),
           vessel_id = col_chr(cmb, "vesselId"),
           gear      = col_chr(cmb, "geartypes_name"),
           ship      = col_chr(cmb, "shiptypes_name"),
           year_from = col_chr(cmb, "shiptypes_yearFrom"),
           year_to   = col_chr(cmb, "shiptypes_yearTo")) |>
      inner_join(idx, by = "index", relationship = "many-to-many") |>
      select(all_of(combined_cols)) |>
      distinct()
  } else {
    read_ev("", combined_cols)
  }

  list(registry = registry, combined = combined)
}
```

```{r}
id_lookup <- bind_rows(vp, afe) |>
  filter(!is.na(MMSI), !is.na(`Vessel ID`)) |>
  distinct(vessel_id = `Vessel ID`, MMSI)

queried <- read_ev(ev_paths$queried, c("vessel_id", "queried_on", "status"))

to_query <- id_lookup |>
  filter(MMSI %in% candidates$MMSI,
         !vessel_id %in% queried$vessel_id[queried$status %in% c("ok", "not_found")]) |>
  pull(vessel_id) |>
  unique()

message(length(to_query), " candidate Vessel IDs not yet queried")

if (run_gfw_api && length(to_query) > 0) {
  for (ids in split(to_query, ceiling(seq_along(to_query) / api_batch_size))) {
    res <- tryCatch(
      gfw_vessel_info(search_type = "id", ids = ids,
                      key = Sys.getenv("GFW_TOKEN"), quiet = TRUE),
      error = function(e) e)

    status <- if (inherits(res, "error")) "error" else if (is.null(res)) "not_found" else "ok"

    if (status == "ok") {
      parsed <- parse_identity(res)
      append_csv(parsed$registry, ev_paths$registry)
      append_csv(parsed$combined, ev_paths$combined)
    }

    append_csv(tibble(vessel_id = ids, queried_on = as.character(Sys.Date()),
                      status = status),
               ev_paths$queried)
    Sys.sleep(0.5)
  }
}
```

```{r}
# summarise identity evidence per MMSI
queried  <- read_ev(ev_paths$queried, c("vessel_id", "queried_on", "status"))
registry <- read_ev(ev_paths$registry, registry_cols)
combined <- read_ev(ev_paths$combined, combined_cols)

queried_mmsi <- id_lookup |>
  filter(vessel_id %in% queried$vessel_id[queried$status %in% c("ok", "not_found")]) |>
  distinct(MMSI)

has_fishing_gear <- function(g) {
  map_lgl(str_split(coalesce(g, ""), fixed("|")), \(x) any(x %in% fishing_gear_codes))
}

gfw_evidence <- queried_mmsi |>
  left_join(registry |>
              group_by(MMSI) |>
              summarise(reg_records = n(),
                        reg_sources = paste(sort(unique(source)), collapse = "|"),
                        reg_gear    = paste(sort(unique(reg_gear)), collapse = "|"),
                        .groups = "drop"),
            by = "MMSI") |>
  left_join(combined |>
              group_by(MMSI) |>
              summarise(gfw_combined = paste(unique(paste0(ship, " ", year_from, "-", year_to)),
                                             collapse = "; "),
                        .groups = "drop"),
            by = "MMSI") |>
  mutate(reg_records    = coalesce(reg_records, 0L),
         reg_fishing    = reg_records > 0 & has_fishing_gear(reg_gear),
         reg_nonfishing = reg_records > 0 & !reg_fishing)

count(gfw_evidence, reg_records > 0, reg_fishing)
```

### External registries

GARFO permit data, USCG vessel documentation and similar sources each come in
their own format. Convert each download to a CSV in
`data/raw/vessel_evidence/registries/` with these columns, then re-run:

| Column | Content |
|---|---|
| `source` | e.g. `GARFO_permits_2024`, `USCG_documentation` |
| `mmsi` | MMSI if the source has it, else blank |
| `call_sign` | Call sign |
| `vessel_name` | Name (shown for review only; never matched on) |
| `is_fishing` | `TRUE` if the source records a fishing permit or fishing service |
| `detail` | Permit category, service type or gear |

Matches use MMSI where the source has it, otherwise the call sign. Names are
never used to match, for the reasons in `00b-plan-revisions.qmd`.

```{r}
norm_call_sign <- function(x) str_remove_all(toupper(coalesce(x, "")), "[^A-Z0-9]")

reg_files <- list.files(file.path(evid_dir, "registries"),
                        pattern = "\\.csv$", full.names = TRUE)

external <- reg_files |>
  map(\(f) read_csv(f, col_types = cols(.default = col_character()), na = c("", "NA"))) |>
  list_rbind()

if (nrow(external) > 0) {
  stopifnot("registry CSVs need source, mmsi, call_sign, vessel_name, is_fishing, detail" =
              all(c("source", "mmsi", "call_sign", "vessel_name", "is_fishing", "detail")
                  %in% names(external)))

  external <- external |> mutate(is_fishing = as.logical(is_fishing),
                                 cs_key = norm_call_sign(call_sign))

  cand_keys <- candidates |>
    distinct(MMSI, call_sign) |>
    mutate(cs_key = norm_call_sign(call_sign))

  ext_matches <- bind_rows(
    cand_keys |> inner_join(filter(external, !is.na(mmsi)),
                            by = c("MMSI" = "mmsi"), suffix = c("", ".ext")),
    cand_keys |> filter(cs_key != "") |>
      inner_join(filter(external, is.na(mmsi)), by = "cs_key", suffix = c("", ".ext"))
  )

  ext_evidence <- ext_matches |>
    group_by(MMSI) |>
    summarise(ext_sources    = paste(sort(unique(source)), collapse = "|"),
              ext_detail     = paste(unique(na.omit(detail)), collapse = "; "),
              ext_fishing    = any(is_fishing %in% TRUE),
              ext_nonfishing = all(is_fishing %in% FALSE),
              .groups = "drop")
} else {
  ext_evidence <- tibble(MMSI = character(), ext_sources = character(),
                         ext_detail = character(), ext_fishing = logical(),
                         ext_nonfishing = logical())
  message("No external registry CSVs in ", file.path(evid_dir, "registries"))
}
```

## Phase 4: Review and decide

### Suggested classifications

| Evidence | Suggestion |
|---|---|
| Registry or permit lists fishing (list 1) | `fishing`, `confirmed` -- review these first |
| Registry or permit lists only non-fishing (list 1) | resolved by evidence; GFW's label stands, no review needed |
| No registry evidence (list 1) | `unknown`, `possible` -- review by hours |
| Charter burst with no prior history (list 2) | exclusion, `probable` |
| Lease-share rise only (list 2) | exclusion, `possible` |
| Registry or permit lists non-fishing (list 3) | `non_fishing`, `confirmed` |
| Name or behaviour only (list 3) | `non_fishing`, `possible` |
| Battery-level call sign (list 4) | `gear_marker`, `probable` |
| MMSI not 9 digits only (list 4) | `unknown`, `possible` -- check whether it is a vessel |

Charter windows are suggested from the first and last Stage 3 day with lease
activity; replace them with Ørsted's dates where available.

```{r}
review_cols <- c("list", "priority", "reason", "MMSI", "names", "flag", "call_sign",
                 "gfw_vessel_type", "gfw_gear_type", "vp_hours", "s1", "s2",
                 "s3", "afe_hours", "h_day", "lease_share", "lease_share_s12",
                 "lease_share_s3", "reg_records", "reg_sources", "reg_gear",
                 "gfw_combined", "ext_sources", "ext_detail",
                 "resolved_by_evidence",
                 "decision", "final_hull_class", "final_gear_class",
                 "final_activity_excluded", "final_valid_from", "final_valid_to",
                 "final_Confidence", "final_Identified_By", "final_evidence_ref",
                 "final_Notes")

suggestions <- candidates |>
  left_join(gfw_evidence, by = "MMSI") |>
  left_join(ext_evidence, by = "MMSI") |>
  mutate(
    across(c(reg_fishing, reg_nonfishing, ext_fishing, ext_nonfishing),
           \(x) coalesce(x, FALSE)),
    fishing_evidence    = reg_fishing | ext_fishing,
    nonfishing_evidence = reg_nonfishing | ext_nonfishing,
    resolved_by_evidence = list == "1_missed_fishing" &
                           nonfishing_evidence & !fishing_evidence,
    decision = NA_character_,
    final_hull_class = case_when(
      list == "4_gear_marker" & str_detect(reason, "battery")    ~ "gear_marker",
      list == "4_gear_marker"                                    ~ "unknown",
      list == "1_missed_fishing" & fishing_evidence              ~ "fishing",
      list == "1_missed_fishing"                                 ~ "unknown",
      list == "2_charter_activity"                               ~ "fishing",
      list == "3_nonfishing_typed_fishing"                       ~ "non_fishing"),
    final_gear_class = NA_character_,
    final_activity_excluded = list == "2_charter_activity",
    final_valid_from = if_else(list == "2_charter_activity",
                               as.character(s3_lease_first), NA_character_),
    final_valid_to   = if_else(list == "2_charter_activity",
                               as.character(s3_lease_last), NA_character_),
    final_Confidence = case_when(
      list == "1_missed_fishing" & fishing_evidence              ~ "confirmed",
      list == "2_charter_activity" & burst                       ~ "probable",
      list == "3_nonfishing_typed_fishing" & nonfishing_evidence ~ "confirmed",
      list == "4_gear_marker" & str_detect(reason, "battery")    ~ "probable",
      TRUE                                                       ~ "possible"),
    final_Identified_By = case_when(
      list %in% c("1_missed_fishing", "3_nonfishing_typed_fishing") & (ext_fishing | ext_nonfishing)
        ~ paste0("external_registry:", ext_sources),
      list %in% c("1_missed_fishing", "3_nonfishing_typed_fishing") & (reg_fishing | reg_nonfishing)
        ~ "gfw_registry",
      list == "2_charter_activity" ~ "behavioural_screen",
      list == "4_gear_marker"      ~ "mmsi_format_screen",
      TRUE                         ~ "screen"),
    final_evidence_ref = NA_character_,
    final_Notes = reason) |>
  # review order within each list: fishing evidence first, evidence-resolved
  # last, then by hours
  arrange(list, desc(fishing_evidence), resolved_by_evidence, desc(vp_hours)) |>
  group_by(list) |>
  mutate(priority = row_number()) |>
  ungroup() |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(any_of(review_cols))
```

### Review file

The review file is merged, not overwritten. Rows with a decision keep it and
their `final_*` edits; undecided rows are refreshed so new evidence (e.g. an
API pass) updates their suggestions; new candidates are added.

```{r}
dir.create(paths$review, recursive = TRUE, showWarnings = FALSE)
review_path <- file.path(paths$review, "vessel_review.csv")

existing <- if (file.exists(review_path)) {
  read_csv(review_path, col_types = cols(.default = col_character()), na = c("", "NA"))
} else {
  NULL
}

decided_cols <- c("decision", str_subset(review_cols, "^final_"))
fresh <- mutate(suggestions, across(everything(), as.character))

review <- if (is.null(existing)) {
  fresh
} else {
  kept <- filter(existing, !is.na(decision))
  bind_rows(
    # new or still-undecided candidates: fresh suggestions
    anti_join(fresh, kept, by = c("list", "MMSI")),
    # decided candidates still screened: fresh evidence, kept decisions
    fresh |>
      select(-all_of(decided_cols)) |>
      inner_join(select(kept, list, MMSI, all_of(decided_cols)), by = c("list", "MMSI")),
    # decided vessels no longer screened (e.g. now in the crosswalk): kept as history
    anti_join(kept, fresh, by = c("list", "MMSI")))
}

review <- arrange(review, list, as.integer(priority))

write_csv(review, review_path, na = "")

stopifnot("decision must be blank, accept or reject" =
            all(review$decision %in% c(NA, "accept", "reject")))
```

### Apply accepted decisions to the crosswalk

```{r}
accepted <- review |>
  filter(decision == "accept") |>
  transmute(MMSI,
            valid_from = final_valid_from,
            valid_to   = final_valid_to,
            Vessel_Name = names,
            hull_class = final_hull_class,
            gear_class = final_gear_class,
            activity_excluded = final_activity_excluded,
            Operation = NA_character_,
            Identified_By = final_Identified_By,
            Confidence = final_Confidence,
            evidence_ref = final_evidence_ref,
            gfw_vessel_type, gfw_gear_type,
            pull_date = as.character(pull_date),
            Notes = final_Notes)

cw_raw <- read_csv(paths$crosswalk, col_types = cols(.default = col_character()),
                   na = c("", "NA"))

new_rows <- accepted |>
  anti_join(cw_raw, by = c("MMSI", "valid_from", "valid_to", "activity_excluded"))

if (nrow(new_rows) > 0) {
  updated <- bind_rows(cw_raw, new_rows) |> select(all_of(cw_cols))

  tmp <- tempfile(fileext = ".csv")
  write_csv(updated, tmp, na = "")
  read_crosswalk(tmp) # validate before replacing the real file

  write_csv(updated, paths$crosswalk, na = "")
  message("Added ", nrow(new_rows), " accepted rows to ", paths$crosswalk)
}

cw <- read_crosswalk(paths$crosswalk)
```

### Review coverage

Reviewing stops for a list when the hours still unreviewed fall below
`screen$review_cutoff` of the dataset's total hours. List 1 vessels whose
registry records only non-fishing gear count as resolved by evidence, so list 1
is only reachable after the API pass and registry matching.

```{r}
total_vp  <- sum(vp$hours)
total_afe <- sum(afe$hours)

review |>
  mutate(vp_hours = as.numeric(vp_hours), afe_hours = as.numeric(afe_hours),
         reviewed = !is.na(decision) | resolved_by_evidence %in% "TRUE") |>
  group_by(list) |>
  summarise(vessels = n(),
            reviewed = sum(reviewed),
            unreviewed_vp_share  = sum(vp_hours[!reviewed]) / total_vp,
            unreviewed_afe_share = sum(afe_hours[!reviewed]) / total_afe,
            .groups = "drop") |>
  mutate(below_cutoff = unreviewed_vp_share < screen$review_cutoff &
                        unreviewed_afe_share < screen$review_cutoff)
```

## Phase 5: Classify every record

Each record is matched to at most one crosswalk row by MMSI and date. Records
with no row keep GFW's label. Three analysis sets follow:

| Set | Fishing hulls GFW missed that are added | Non-fishing hulls and charter activity removed |
|---|---|---|
| **Strict** | none | `confirmed` only |
| **Primary** | `confirmed` | `confirmed` and `probable` |
| **Inclusive** | `confirmed`, `probable`, `possible` | `confirmed` and `probable` |

Gear markers are removed from every set. The sets differ along two axes, so
they are not strictly nested: Strict adds nothing but also removes the least.
`possible` exclusions (e.g. ESS Pride) are removed from no set; list them as
a stated sensitivity rather than as a filter.

```{r}
classify <- function(df, cw) {
  cw_join <- cw |>
    transmute(MMSI,
              cw_from = coalesce(valid_from, lo_date),
              cw_to   = coalesce(valid_to, hi_date),
              cw_hull = hull_class,
              cw_gear = gear_class,
              cw_excluded = activity_excluded,
              cw_confidence = Confidence,
              cw_identified_by = Identified_By)

  out <- df |>
    left_join(cw_join, by = join_by(MMSI, between(date, cw_from, cw_to)))

  stopifnot("a record matched more than one crosswalk row" = nrow(out) == nrow(df))

  out |>
    mutate(
      gfw_fishing = `Vessel Type` %in% "FISHING",
      conf_cp     = cw_confidence %in% c("confirmed", "probable"),
      is_marker   = cw_hull %in% "gear_marker",
      nonfish_c   = cw_hull %in% "non_fishing" & cw_confidence %in% "confirmed",
      nonfish_cp  = cw_hull %in% "non_fishing" & conf_cp,
      added_c     = cw_hull %in% "fishing" & !gfw_fishing & cw_confidence %in% "confirmed",
      added_any   = cw_hull %in% "fishing" & !gfw_fishing,
      excl_c      = cw_excluded %in% TRUE & cw_confidence %in% "confirmed",
      excl_cp     = cw_excluded %in% TRUE & conf_cp,
      in_strict    = gfw_fishing & !is_marker & !nonfish_c & !excl_c,
      in_primary   = (gfw_fishing | added_c) & !is_marker & !nonfish_cp & !excl_cp,
      in_inclusive = (gfw_fishing | added_any) & !is_marker & !nonfish_cp & !excl_cp,
      gear_class  = coalesce(cw_gear, gear_class_of(`Gear Type`))) |>
    select(-cw_from, -cw_to, -conf_cp)
}

vp_cls  <- classify(vp, cw)
afe_cls <- classify(afe, cw)
```

```{r}
set_cols <- c("in_strict", "in_primary", "in_inclusive")

bind_rows(VP = vp_cls, AFE = afe_cls, .id = "dataset") |>
  group_by(dataset) |>
  summarise(across(all_of(set_cols), \(x) sum(hours[x])),
            total = sum(hours), .groups = "drop")
```

```{r}
write_csv(vp_cls,  file.path(paths$out_gfw, "gfw_vp_id_classified.csv"),  na = "")
write_csv(afe_cls, file.path(paths$out_gfw, "gfw_afe_id_classified.csv"), na = "")
```

## Phase 6: Account for what is left

### Stage totals and Stage 2 → 3 change by set

```{r}
to_long <- function(df) {
  df |>
    pivot_longer(all_of(set_cols), names_to = "set", values_to = "keep") |>
    filter(keep) |>
    mutate(set = factor(str_remove(set, "in_"),
                        levels = c("strict", "primary", "inclusive")))
}

stage_by_set <- bind_rows(VP = to_long(vp_cls), AFE = to_long(afe_cls), .id = "dataset") |>
  group_by(dataset, set, `Development Stage`) |>
  summarise(hours = sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop")

stage_by_set

stage_by_set |>
  select(-n_mmsi) |>
  pivot_wider(names_from = `Development Stage`, values_from = hours) |>
  mutate(change_s2_s3_pct = round(100 * (`Stage 3` - `Stage 2`) / `Stage 2`, 1))
```

### Fishing effort missing for added vessels

GFW computes AFE only for vessels it treats as fishing, so hulls added in
Primary or Inclusive have presence hours but almost no effort hours. Report
this as a known undercount. The imputation below applies each stage's
effort-to-presence ratio (from Strict-set vessels, by gear class) to the added
vessels' presence hours; use it as a sensitivity check only.

```{r}
added <- vp_cls |> filter(added_any, in_inclusive)

ratio <- inner_join(
  vp_cls  |> filter(in_strict) |> group_by(`Development Stage`, gear_class) |>
    summarise(vp = sum(hours), .groups = "drop"),
  afe_cls |> filter(in_strict) |> group_by(`Development Stage`, gear_class) |>
    summarise(afe = sum(hours), .groups = "drop"),
  by = c("Development Stage", "gear_class")) |>
  mutate(afe_per_vp = afe / vp)

added |>
  group_by(`Development Stage`, gear_class,
           set = if_else(added_c, "primary + inclusive", "inclusive only")) |>
  summarise(vp_hours = sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop") |>
  left_join(afe_cls |> filter(added_any) |>
              group_by(`Development Stage`, gear_class) |>
              summarise(afe_hours_reported = sum(hours), .groups = "drop"),
            by = c("Development Stage", "gear_class")) |>
  left_join(select(ratio, `Development Stage`, gear_class, afe_per_vp),
            by = c("Development Stage", "gear_class")) |>
  mutate(afe_hours_reported = coalesce(afe_hours_reported, 0),
         afe_hours_imputed  = vp_hours * afe_per_vp)
```

### Cell × stage tables by set

Same zero-filled layout as `gfw_fishnet_grid.qmd`, with a `set` column, ready
for the spatial analysis.

```{r}
cell_stage_sets <- function(df, value) {
  to_long(df) |>
    group_by(cell_id, `Development Stage`, set) |>
    summarise("{value}" := sum(hours), n_mmsi = n_distinct(MMSI), .groups = "drop") |>
    complete(cell_id = grid$cell_id, `Development Stage`, set,
             fill = set_names(list(0, 0L), c(value, "n_mmsi")))
}

vp_cell_stage_sets  <- cell_stage_sets(vp_cls,  "vp_hours")
afe_cell_stage_sets <- cell_stage_sets(afe_cls, "afe_hours")

stopifnot(nrow(vp_cell_stage_sets) == nrow(grid) * 3 * 3)

write_csv(vp_cell_stage_sets,  file.path(paths$out_grid, "gfw_vp_cell_stage_sets.csv"))
write_csv(afe_cell_stage_sets, file.path(paths$out_grid, "gfw_afe_cell_stage_sets.csv"))
```

## Phase 7: Repeatability checks

### GFW label drift

GFW's labels change between pulls (CYGNUS was typed FISHING/INCONCLUSIVE in an
earlier pull and CARGO in this one). Any crosswalk vessel whose current label
differs from the one recorded when its row was written is listed here for a
second look.

```{r}
cw |>
  distinct(MMSI, Vessel_Name, gfw_vessel_type, gfw_gear_type, pull_date) |>
  left_join(gfw_labels, by = "MMSI", suffix = c("_recorded", "_current")) |>
  filter(is.na(gfw_vessel_type_current) |
           gfw_vessel_type_recorded != gfw_vessel_type_current |
           gfw_gear_type_recorded != gfw_gear_type_current)
```

### Crosswalk summary for the methods

```{r}
touched <- function(df) {
  df |>
    filter(!is.na(cw_hull)) |>
    group_by(cw_hull, cw_excluded, cw_confidence) |>
    summarise(hours = sum(hours), .groups = "drop")
}

count(cw, hull_class, activity_excluded, Confidence, name = "rows")

full_join(touched(vp_cls), touched(afe_cls),
          by = c("cw_hull", "cw_excluded", "cw_confidence"),
          suffix = c("_vp", "_afe")) |>
  mutate(share_vp  = round(hours_vp  / total_vp,  4),
         share_afe = round(hours_afe / total_afe, 4))
```
