---
title: "MM GFW 0.01° Fishnet Grid (UTM 19N)"
author: "Malachy McCaffrey"
date: today
format: html
params:
  fishnet: ""     # override the grid output folder (paths$fishnet)
  interim_gfw: "" # override the aggregate-table output folder (paths$interim_gfw)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# run from the repo root (the folder holding renv.lock) so the relative paths
# in r/R/paths.R resolve whether the qmd is rendered or run chunk by chunk
root <- getwd()
while (!file.exists(file.path(root, "renv.lock")) && dirname(root) != root) root <- dirname(root)
knitr::opts_knit$set(root.dir = root)
```

## Objective

Reproduce Global Fishing Watch's 0.01° grid for the vessel presence (VP) and
apparent fishing effort (AFE) records pulled in `gfw_vp_afe_datapull.qmd`,
reproject it to WGS 84 / UTM zone 19N (EPSG:32619), and attach the data to it
for aggregation and spatial analysis.

### Data status: aggregates describe the raw GFW data

The aggregate tables built here describe the **raw, unfiltered** GFW records in
`data/raw/gfw/gfw_vp_id.csv` and `gfw_afe_id.csv`. They include every vessel
type and apply no vessel removals or fishing-hour threshold. They are **not**
fishing-vessel totals and should not be analysed as such.

The data have not yet been re-processed because GFW's vessel classifications
produce both false negatives and false positives:

- **False negatives:** fishing vessels typed as `OTHER`, `GEAR`, or
  `PASSENGER` (e.g. FV JULIE G), so filtering on `Vessel Type == "FISHING"`
  drops real fishing activity.
- **False positives:** non-fishing vessels typed as `FISHING` (e.g. CYGNUS), so
  the same filter keeps non-fishing activity.

Both problems affect VP and AFE; `gfw_vessel_filtering.qmd` addresses them.
The grid itself does not depend on that work, because it is built from the
cells present in the raw VP data.

## Plan

1. **Snap coordinates to integer cell indices.** GFW `Lat`/`Lon` mark the
   **lower-left corner** of each 0.01° cell. Values carry float32 noise
   (e.g. `-71.209999`), so snap with `round(x / 0.01)` -- never `floor()`, which
   drops values like `41.0699999` into the wrong cell. The integer pair
   `cell_id = "ix_iy"` is the join key from here on; floating-point
   coordinates are never compared. `read_gfw()` in `r/R/gfw_helpers.R` does
   this for every script.
2. **Grid = the distinct cells in `gfw_vp_id`.** Every AFE cell must already be
   among them.
3. **Build each cell from its lower-left corner** in WGS 84, spanning
   `Lon`–`Lon + 0.01` and `Lat`–`Lat + 0.01`. Corners come from integer
   arithmetic so adjacent cells share bit-identical vertices.
4. **Reproject the polygons to EPSG:32619.** Each cell becomes a parallelogram
   of roughly 840 m E–W by 1,110 m N–S, rotated by grid convergence. Do not
   build a new square UTM fishnet or resample a raster, either of which would
   redistribute hours between cells.
5. **Add cell attributes:** `area_km2` (from UTM), `aoi_frac`, and `aoi_class`
   (inside / edge / outside).
6. **Grid-cell decisions** (see below): keep data cells outside the AOI;
   exclude AOI-touching cells that have no records.
7. **Attach the raw data with a table join on `cell_id`**, aggregate per
   cell × stage and per cell × month, and fill zeros across the grid. These
   are raw-data aggregates (all vessel types); see *Data status*.
8. **Neighbours** from queen contiguity on the UTM grid.

### Grid-cell decisions

Under the lower-left anchor, two sets of cells sit at the AOI boundary:

- **Data cells outside the AOI -- kept.** They lie along the north and east
  edges. They are cells GFW returned for the AOI query and hold real records,
  so they are flagged `aoi_class == "outside"` and can be dropped in a
  sensitivity check.
- **AOI-touching cells with no records -- excluded.** They lie along the south
  and west edges. No VP or AFE record has been returned under their keys, so
  they are not part of the grid and are not zero-filled.

The mirror-image pattern (extra cells to the NE, missing cells to the SW) is
consistent with GFW selecting cells for a `USER_SHAPEFILE` region by a rule
that differs from the lower-left labelling. The counts for this pull are
reported under *Grid-cell decision checks*; diagnostics are in
`r/scratchpad/gridcell_decision.R`.

### Why a table join, not a spatial join

Each record already names its cell, so the join only has to read the label.
With lower-left anchoring every point sits exactly on a cell corner shared by
four cells, and float noise nudges it to an arbitrary side of the edge. On a
20,000-row sample of the 2026-10-06 `gfw_vp_id` pull:

- `st_intersects` (WGS 84) matched most points to 2 or 4 cells, turning 20,000
  rows into 44,183 and inflating VP hours 2.19×.
- `st_within` (WGS 84) matched none of the 20,000 points.
- `st_intersects` after reprojecting to UTM put 24,197 rows in the wrong cell.

None of these fail loudly. The table join is exact, one-to-one, checkable with
`anti_join()`, and identical in any language. Spatial joins are reserved for
linking the grid to layers that do not share the key (lease areas, turbines,
the AOI), done once at the cell level as in the `aoi_frac` step.

## Libraries and shared code

```{r}
library(tidyverse)
library(sf)
library(spdep)

source("r/R/paths.R")
source("r/R/gfw_helpers.R")

paths <- gfw_paths_with(if (exists("params")) params else list())
```

## Parameters

```{r}
crs_utm <- 32619 # WGS 84 / UTM zone 19N
```

## Read raw data and snap to cell indices

```{r}
vp  <- read_gfw(paths$raw_vp)
afe <- read_gfw(paths$raw_afe)

# every coordinate must sit on the 0.01 lattice
snap_err <- with(bind_rows(vp, afe), max(abs(Lon / 0.01 - ix), abs(Lat / 0.01 - iy)))
stopifnot("coordinates are not on the 0.01 degree lattice" = snap_err < 0.01)

# every record needs a development stage (r/R/gfw_helpers.R)
stopifnot("records with no development stage" =
            !anyNA(vp$`Development Stage`) && !anyNA(afe$`Development Stage`))
```

## Grid cells from the VP data

```{r}
cells <- distinct(vp, cell_id, ix, iy)

stopifnot("AFE has cells absent from the VP grid" =
            all(afe$cell_id %in% cells$cell_id))
```

## Build cell polygons from the lower-left corner

```{r}
# Corners come from integer arithmetic so adjacent cells share bit-identical
# vertices, which keeps topology exact after reprojection.
cell_poly <- function(ix, iy) {
  x <- (ix + c(0L, 1L, 1L, 0L, 0L)) / 100
  y <- (iy + c(0L, 0L, 1L, 1L, 0L)) / 100
  st_polygon(list(cbind(x, y)))
}

grid_wgs <- st_sf(cells,
                  geometry = st_sfc(map2(cells$ix, cells$iy, cell_poly),
                                    crs = 4326))
```

## Reproject and add cell attributes

```{r}
aoi_utm <- st_read(paths$aoi, quiet = TRUE) |>
  st_union() |>
  st_transform(crs_utm)

grid_utm <- st_transform(grid_wgs, crs_utm) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6)

# fraction of each cell inside the AOI; cells outside it get 0
in_aoi <- st_intersection(select(grid_utm, cell_id), aoi_utm)

grid_utm <- grid_utm |>
  left_join(tibble(cell_id = in_aoi$cell_id,
                   aoi_m2  = as.numeric(st_area(in_aoi))),
            by = "cell_id") |>
  mutate(aoi_frac  = coalesce(aoi_m2 / (area_km2 * 1e6), 0),
         aoi_class = factor(case_when(aoi_frac == 0   ~ "outside",
                                      aoi_frac < 0.99 ~ "edge",
                                      TRUE            ~ "inside"),
                            levels = c("inside", "edge", "outside"))) |>
  select(-aoi_m2)

count(st_drop_geometry(grid_utm), aoi_class)
summary(grid_utm$area_km2)
```

## Grid-cell decision checks

```{r}
# Excluded: AOI-touching cells with no records are not added to the grid.
# Built here only to document the decision.
lattice <- expand_grid(ix = seq(min(cells$ix) - 1L, max(cells$ix) + 1L),
                       iy = seq(min(cells$iy) - 1L, max(cells$iy) + 1L)) |>
  mutate(cell_id = sprintf("%d_%d", ix, iy)) |>
  filter(!cell_id %in% cells$cell_id)

empty_utm <- st_sf(lattice,
                   geometry = st_sfc(map2(lattice$ix, lattice$iy, cell_poly),
                                     crs = 4326)) |>
  st_transform(crs_utm)

empty_utm <- empty_utm[lengths(st_intersects(empty_utm, aoi_utm)) > 0, ]

stopifnot("excluded empty cells leaked into the grid" =
            !any(empty_utm$cell_id %in% grid_utm$cell_id))

outside_ids <- grid_utm$cell_id[grid_utm$aoi_class == "outside"]
```

The grid has `r nrow(grid_utm)` cells. `r length(outside_ids)` lie entirely
outside the AOI and are kept; they hold `r sum(vp$cell_id %in% outside_ids)`
VP records (`r round(100 * sum(vp$hours[vp$cell_id %in% outside_ids]) / sum(vp$hours), 2)`%
of VP hours). `r nrow(empty_utm)` AOI-touching cells have no records and are
excluded.

## Map check

```{r}
ggplot() +
  geom_sf(data = grid_utm, aes(fill = aoi_class), color = "white", linewidth = 0.05) +
  geom_sf(data = empty_utm, fill = NA, color = "black", linewidth = 0.3, linetype = "22") +
  geom_sf(data = aoi_utm, fill = NA, color = "black", linewidth = 0.6) +
  scale_fill_manual(values = c(inside = "#d9d8d2", edge = "#2a78d6", outside = "#eb6834")) +
  labs(fill = NULL,
       caption = "Dashed: AOI-touching cells with no records (excluded).") +
  theme_minimal()
```

## Attach raw data with a table join and aggregate

These aggregates describe the **raw GFW data** -- all vessel types, no vessel
removals, no fishing-hour threshold. They are not fishing-vessel totals; see
*Data status* above. `Year Month` and `Development Stage` come from
`read_gfw()`, so the stage boundaries are the ones in `r/R/gfw_helpers.R`.

```{r}
# every record must land in a grid cell -- both should return zero rows
grid_ids <- st_drop_geometry(grid_utm) |> select(cell_id)

stopifnot(
  "VP records with no grid cell"  = nrow(anti_join(vp,  grid_ids, by = "cell_id")) == 0,
  "AFE records with no grid cell" = nrow(anti_join(afe, grid_ids, by = "cell_id")) == 0
)

count(vp, `Development Stage`)
```

```{r}
# cell x development stage, zero-filled across the grid
# counting unit is MMSI, not Vessel ID (see references/data_dictionary.md)

vp_cell_stage <- vp |>
  group_by(cell_id, `Development Stage`) |>
  summarise(vp_hours = sum(hours),
            n_mmsi   = n_distinct(MMSI),
            .groups  = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Development Stage`,
           fill = list(vp_hours = 0, n_mmsi = 0L))

afe_cell_stage <- afe |>
  group_by(cell_id, `Development Stage`) |>
  summarise(afe_hours = sum(hours),
            n_mmsi    = n_distinct(MMSI),
            .groups   = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Development Stage`,
           fill = list(afe_hours = 0, n_mmsi = 0L))
```

```{r}
# cell x month, zero-filled across the grid and every month in the record

months <- seq(min(vp$`Year Month`), max(vp$`Year Month`), by = "month")

vp_cell_month <- vp |>
  group_by(cell_id, `Year Month`) |>
  summarise(vp_hours = sum(hours),
            n_mmsi   = n_distinct(MMSI),
            .groups  = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Year Month` = months,
           fill = list(vp_hours = 0, n_mmsi = 0L))

afe_cell_month <- afe |>
  group_by(cell_id, `Year Month`) |>
  summarise(afe_hours = sum(hours),
            n_mmsi    = n_distinct(MMSI),
            .groups   = "drop") |>
  complete(cell_id = grid_utm$cell_id, `Year Month` = months,
           fill = list(afe_hours = 0, n_mmsi = 0L))

stopifnot(nrow(vp_cell_month) == nrow(grid_utm) * length(months))
```

```{r}
# spatial versions with area-normalised density, since cell area varies

vp_stage_sf <- grid_utm |>
  left_join(vp_cell_stage, by = "cell_id") |>
  mutate(vp_hours_km2 = vp_hours / area_km2)

afe_stage_sf <- grid_utm |>
  left_join(afe_cell_stage, by = "cell_id") |>
  mutate(afe_hours_km2 = afe_hours / area_km2)
```

## Neighbours

```{r}
# Cells are taller (N-S) than wide (E-W): a distance band would link E-W
# neighbours but skip N-S ones, so use contiguity.
nb <- poly2nb(grid_utm, queen = TRUE, snap = 0)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

summary(card(nb))
stopifnot("grid has cells with no neighbours" = all(card(nb) > 0))
```

## Export

```{r}
# grid (supersedes archive/fishnet/Orsted_sqGrid_utm19n.shp, which was built
# treating Lat/Lon as centroids)
st_write(grid_utm, file.path(paths$fishnet, basename(paths$grid)), delete_dsn = TRUE)

# raw-data aggregates (all vessel types); file names keep the raw "_id" suffix
# so they are not mistaken for fishing-vessel totals
write_csv(vp_cell_stage,  file.path(paths$interim_gfw, "gfw_vp_id_cell_stage.csv"))
write_csv(afe_cell_stage, file.path(paths$interim_gfw, "gfw_afe_id_cell_stage.csv"))
write_csv(vp_cell_month,  file.path(paths$interim_gfw, "gfw_vp_id_cell_month.csv"))
write_csv(afe_cell_month, file.path(paths$interim_gfw, "gfw_afe_id_cell_month.csv"))
```
